---
title: "Analyzing the eLife Fitness Dataset with MPCurver"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Analyzing the eLife Fitness Dataset with MPCurver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5.5
)
```

## Overview

This vignette shows how to use the public `fit_mpcurve()` interface on the
eLife fitness dataset. The goal is not just to produce a fit, but to explain
the modeling logic step by step.

We will use `MPCurver` to answer three practical questions:

1. Can a single latent ordering explain mutant fitness across environments?
2. If we encourage smoother trajectories, do we keep the same ordering while
   improving interpretability?
3. What feature partition and conditional trajectories appear under a
   scientifically chosen fixed two-ordering model, with and without known
   measurement error?

The recommended workflow throughout is the default `cavi` backend exposed by
`fit_mpcurve()`.

## Load package and data

```{r load}
if (requireNamespace("MPCurver", quietly = TRUE)) {
  library(MPCurver)
} else if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(".", quiet = TRUE)
} else {
  stop("Need either an installed MPCurver package or devtools::load_all().")
}

set.seed(42)

locate_elife_csv <- function() {
  pkg_path <- system.file("extdata", "elife-61271-fig2-data1-v2.csv", package = "MPCurver")
  if (nzchar(pkg_path) && file.exists(pkg_path)) {
    return(pkg_path)
  }

  src_path <- file.path("inst", "extdata", "elife-61271-fig2-data1-v2.csv")
  if (file.exists(src_path)) {
    return(src_path)
  }

  stop("Could not locate elife-61271-fig2-data1-v2.csv.")
}

csv_path <- locate_elife_csv()
dat <- read.csv(csv_path, check.names = FALSE)

fitness_cols <- grep("_fitness$", names(dat), value = TRUE)
error_cols <- grep("_error$", names(dat), value = TRUE)

X <- as.matrix(dat[, fitness_cols, drop = FALSE])
mode(X) <- "numeric"
colnames(X) <- sub("_fitness$", "", fitness_cols)

S <- as.matrix(dat[, error_cols, drop = FALSE])
mode(S) <- "numeric"
colnames(S) <- sub("_error$", "", error_cols)

meta <- dat[, c("barcode", "gene", "type", "ploidy", "class", "mutation_type"), drop = FALSE]
```


Just using the error columns as the true SEs might be too risky, so we will adjust the SEs using t-statistics-adjustment:

```{r adjust-se}
t_adjust_se <- function(Y, S, nu, tol = 1e-12) {
  if (!is.matrix(Y) || !is.numeric(Y)) {
    stop("`Y` must be a numeric matrix.")
  }
  if (!is.matrix(S) || !is.numeric(S)) {
    stop("`S` must be a numeric matrix.")
  }
  if (!all(dim(Y) == dim(S))) {
    stop("`Y` and `S` must have the same dimensions.")
  }
  if (any(S < 0, na.rm = TRUE)) {
    stop("`S` must be nonnegative.")
  }
  
  if (length(nu) == 1L) {
    if (!is.numeric(nu) || is.na(nu) || nu <= 0) {
      stop("`nu` must be a positive number.")
    }
    nu_mat <- matrix(nu, nrow = nrow(Y), ncol = ncol(Y))
  } else {
    if (!is.matrix(nu) || !is.numeric(nu) || !all(dim(nu) == dim(Y))) {
      stop("`nu` must be either a positive scalar or a numeric matrix with the same dimensions as `Y`.")
    }
    if (any(nu <= 0, na.rm = TRUE)) {
      stop("All entries of `nu` must be positive.")
    }
    nu_mat <- nu
  }
  
  z_t <- Y / S
  
  # Main calculation:
  # S_tilde = S * z_t / qnorm(pt(z_t, df = nu))
  p_t <- pt(z_t, df = nu_mat)
  z_n <- qnorm(p_t)
  
  S_tilde <- S * z_t / z_n
  
  # Numerical fix for z_t close to 0:
  # limit of z / qnorm(pt(z, df=nu)) as z -> 0 is dnorm(0) / dt(0, df=nu)
  near_zero <- is.finite(z_t) & abs(z_t) < tol
  if (any(near_zero, na.rm = TRUE)) {
    limit_factor <- dnorm(0) / dt(0, df = nu_mat[near_zero])
    S_tilde[near_zero] <- S[near_zero] * limit_factor
  }
  
  # Preserve zeros
  zero_se <- (S == 0) & !is.na(S)
  if (any(zero_se)) {
    S_tilde[zero_se] <- 0
  }
  
  dimnames(S_tilde) <- dimnames(S)
  S_tilde
}
S <- t_adjust_se(X, S, nu = 4)
```



In this dataset:

- each row is a mutant;
- each column is an environment;
- each matrix entry is a fitness score;
- and the paired `*_error` columns provide observation-level measurement
  standard deviations that we can optionally pass to `fit_mpcurve()` via `S`.

## First look at the data

```{r helpers, include=FALSE}
environment_category <- function(env) {
  if (grepl("Ferm|Day", env)) return("time course")
  if (grepl("NaCl|KCl", env)) return("salt stress")
  if (grepl("Ben|Flu|GdA|DMSO", env)) return("drug / chemical")
  if (grepl("Gluc|Baffle", env)) return("glucose / batch")
  if (grepl("Raf|Gly|EtOH|Suc", env)) return("carbon source")
  "other"
}

env_category <- vapply(colnames(X), environment_category, character(1))

plot_ordered_heatmap <- function(X, order_idx, env_idx = seq_len(ncol(X)),
                                 main = "", zlim = c(-2.5, 2.5)) {
  Xsub <- X[order_idx, env_idx, drop = FALSE]
  Xsub <- scale(Xsub)
  Xsub[!is.finite(Xsub)] <- 0
  pal <- colorRampPalette(c("#313695", "#74add1", "#f7f7f7", "#f46d43", "#a50026"))(101)
  image(
    x = seq_len(nrow(Xsub) + 1L),
    y = seq_len(ncol(Xsub) + 1L),
    z = Xsub,
    col = pal,
    axes = FALSE,
    xlab = "Mutants ordered by inferred pseudotime",
    ylab = "",
    zlim = zlim,
    main = main
  )
  axis(2, at = seq_len(ncol(Xsub)), labels = colnames(Xsub), las = 2, cex.axis = 0.55)
  box()
}

```

```{r data-summary}
data.frame(
  quantity = c("mutants", "environments", "fitness columns", "error columns"),
  value = c(nrow(X), ncol(X), length(fitness_cols), length(error_cols))
) |>
  knitr::kable()

sort(table(env_category), decreasing = TRUE) |>
  as.data.frame() |>
  stats::setNames(c("category", "n_environments")) |>
  knitr::kable()
```

This is a good MPCurver example because the environments span several broad
condition types, while the mutants are shared across all columns. That makes it
natural to ask whether the same latent mutant ordering explains every
environment, or whether different subsets of environments prefer different
orderings.

## Step 1: Fit a single-ordering MPCurve

We start with the simplest question: can one latent ordering explain the main
structure in the dataset?

The call below uses the main user-facing API:

- `method = "isomap"` supplies the initialization;
- `K = 50` uses a 50-bin latent grid;
- `rw_q = 2` uses a second-order random-walk prior for smooth trajectories.

```{r uni-fit}
fit_unidir <- fit_mpcurve(
  X,
  method = "isomap",
  K = 50,
  rw_q = 2,
  verbose = FALSE
)
```

### Check convergence

For the default `cavi` backend, the main quantity to inspect is the ELBO trace.

```{r uni-convergence, out.width = "100%", fig.alt = "ELBO trace for the single-ordering CAVI fit, showing convergence over iterations."}
plot(fit_unidir, plot_type = "elbo")
```

### Inspect a few fitted trajectories

```{r uni-trajectories, fig.height = 8, out.width = "100%", fig.alt = "Four panels showing fitted single-ordering CAVI trajectories for the first four environment dimensions."}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_unidir, dims = 1)
plot(fit_unidir, dims = 2)
plot(fit_unidir, dims = 3)
plot(fit_unidir, dims = 4)
par(mfrow = c(1, 1))
```

The first fit already recovers coherent structure, but the trajectories are
still rougher than we would like for interpretation. 
This is a common pattern on real data: the latent ordering may be sensible even when the fitted curves need stronger regularization.

## Step 2: Encourage smoother trajectories

We now refit the same single-ordering model, but add a stronger prior on
`lambda_j` through `lambda_sd_prior_rate`. Larger values encourage smoother
trajectories.

```{r uni-fit-smooth}
fit_unidir_smooth <- fit_mpcurve(
  X,
  method = "isomap",
  K = 50,
  iter = 500,
  rw_q = 2,
  lambda_sd_prior_rate = 1e4,
  verbose = FALSE
)
```

```{r uni-convergence-smooth, out.width = "100%", fig.alt = "ELBO trace for the smoothed single-ordering CAVI fit."}
plot(fit_unidir_smooth, plot_type = "elbo")
fit_unidir_smooth
```

```{r uni-trajectories-smooth, fig.height = 8, out.width = "100%", fig.alt = "Four panels showing fitted trajectories under the smoothed single-ordering CAVI fit."}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_unidir_smooth, dims = 1)
plot(fit_unidir_smooth, dims = 3)
plot(fit_unidir_smooth, dims = 7)
plot(fit_unidir_smooth, dims = 9)
par(mfrow = c(1, 1))
```

These curves are easier to read: the global ordering remains, but the fitted
environment trajectories are less wiggly.

A useful sanity check is whether smoothing changes the inferred mutant ordering
itself, or mainly changes the trajectory estimates.

```{r compare-single-orderings, out.width = "100%", fig.alt = "Scatter plot comparing inferred mutant pseudotime between the default and smoothed single-ordering fits."}
cor(
  fit_unidir_smooth$locations$mean$pseudotime,
  fit_unidir$locations$mean$pseudotime
)

plot(
  fit_unidir_smooth$locations$mean$pseudotime,
  fit_unidir$locations$mean$pseudotime,
  xlab = "Latent position (smoother fit)",
  ylab = "Latent position (less regularized fit)"
)
abline(0, 1, lty = 2)
```

In this dataset the two fits stay close, which is reassuring: stronger
regularization mostly improves interpretability rather than changing the basic
ordering story.

## Step 3: Explore a fixed two-ordering representation

The single-ordering fit assumes that every environment varies smoothly along
the same latent mutant axis. As a scientifically motivated sensitivity
analysis, we can instead ask how the environments partition under two fixed
latent orderings. This is an exploratory representation, not an automatic test
that $M=2$ is preferable to $M=1$.

To examine that idea, we fit a structural partition model with a scientifically
chosen fixed value $M=2$. The structural ELBO is a fitting objective for a
given $M$; it is not used here as an automatic complexity penalty for choosing
among different values of $M$.

For every `intrinsic_dim >= 2`, the fitted object has one canonical
$q(C)q(Z)q(U\mid Z)$ state. Estimated feature variances form one shared
length-$d$ `params$sigma2` vector across orderings. There are no
active/inactive, freeze, or drop states, and greedy cross-$M$ selection is
currently unavailable.

```{r partition-fit}
fit_partition <- fit_mpcurve(
  X,
  method = "isomap",
  intrinsic_dim = 2,
  ridge = 0,
  K = 50,
  iter = 500,
  rw_q = 2,
  lambda_sd_prior_rate = 1e4,
  verbose = FALSE
)
```

```{r partition-summary}
fit_partition
```

### Inspect conditional structural trajectories

For `plot_type = "mu"`, each ordering panel shows the canonical conditional
posterior mean $E[U_j\mid Z_j=m]$. The title also reports the feature-assignment
probability $q(Z_j=m)$. These are conditional alternatives, so the plotted
trajectory is not multiplied by its assignment probability.

```{r partition-plot, out.width = "100%", fig.alt = "Conditional posterior mean trajectory for environment dimension 1 under each of two fixed orderings."}
plot(fit_partition, plot_type = "mu", dims = 1)
```

### Which environments were assigned to which ordering?

```{r partition-table}
assignment_tbl <- data.frame(
  environment = colnames(X),
  category = env_category,
  ordering = fit_partition$partition$assign,
  stringsAsFactors = FALSE
)

assignment_tbl[order(assignment_tbl$ordering, assignment_tbl$category, assignment_tbl$environment), ] |>
  knitr::kable(row.names = FALSE)
```

### How similar are the inferred orderings?

```{r partition-correlation}
ordering_names <- names(fit_partition$locations)
ordering_names <- ordering_names[ordering_names %in% unique(fit_partition$partition$assign)]

t_df <- setNames(
  as.data.frame(lapply(ordering_names, function(nm) {
    fit_partition$locations[[nm]]$mean$pseudotime
  })),
  ordering_names
)

cor(t_df) |>
  knitr::kable(digits = 3)
```



## Step 4: Refit with known measurement error

The raw file includes paired `*_error` columns, and `fit_mpcurve()` can use
them directly through the `S` argument. In the public API, `S` is interpreted
as a matrix of known measurement standard deviations.

This matters here because ignoring heterogeneous measurement error can make a
fixed multi-ordering fit attribute noisy columns to apparently distinct
trajectory structure.

### Single-ordering fit with known measurement SD

```{r uni-fit-known-sd}
fit_unidir_known_sd <- fit_mpcurve(
  X,
  S = S,
  method = "isomap",
  K = 50,
  iter = 500,
  rw_q = 2,
  lambda_sd_prior_rate = 1e4,
  verbose = FALSE
)
```

```{r uni-known-sd-convergence, out.width = "100%", fig.alt = "ELBO trace for the single-ordering CAVI fit with known measurement standard deviations."}
plot(fit_unidir_known_sd, plot_type = "elbo")
```

```{r uni-known-sd-trajectories, fig.height = 8, out.width = "100%", fig.alt = "Four panels showing fitted trajectories for the single-ordering CAVI fit with known measurement SDs."}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_unidir_known_sd, dims = 1)
plot(fit_unidir_known_sd, dims = 2)
plot(fit_unidir_known_sd, dims = 31)
plot(fit_unidir_known_sd, dims = 45)
par(mfrow = c(1, 1))
```

### Fixed two-ordering fit with known measurement SD

We now refit the same fixed $M=2$ structural model while accounting for
measurement error. This isolates the effect of known noise without mixing it
with a separate dimension-selection procedure.

```{r partition-fit-known-sd}
fit_partition_known_sd <- fit_mpcurve(
  X,
  S = S,
  method = "isomap",
  intrinsic_dim = 2,
  ridge = 0,
  K = 50,
  iter = 500,
  rw_q = 2,
  lambda_sd_prior_rate = 1e4,
  verbose = FALSE
)

fit_partition_known_sd
```

```{r known-sd-partition-trajectories, fig.height = 5.5, out.width = "100%", fig.alt = "Conditional posterior mean trajectory for environment dimension 31 under each of two fixed orderings with known measurement SDs."}
plot(fit_partition_known_sd, plot_type = "mu", dims = 31)
```



## Takeaways

This dataset gives a useful template for real-world `fit_mpcurve()` analyses.

1. Start with a single-ordering fit to establish the basic latent ordering.
2. Use `lambda_sd_prior_rate` when the ordering looks sensible but the fitted
   trajectories are too rough to interpret comfortably.
3. Use a fixed `intrinsic_dim` to study a scientifically plausible number of
   ordering systems; assess different values through stability or external
   validation rather than the raw structural ELBO alone.
4. If measurement error is available, pass it through `S`; otherwise a fixed
   multi-ordering fit may interpret noisy environments as distinct trajectory
   structure.
