---
title: "Analyzing the eLife Fitness Dataset with MPCurver"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Analyzing the eLife Fitness Dataset with MPCurver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5.5
)
```

## Overview

This vignette uses MPCurver to study how mutant fitness varies across
environments. It develops an analysis from a single latent ordering to a
two-ordering feature partition, and examines how smoothing and measurement
uncertainty affect the fitted trajectories.

We will use `MPCurver` to answer three practical questions:

1. Can a single latent ordering explain mutant fitness across environments?
2. If we encourage smoother trajectories, do we keep the same ordering while
   improving interpretability?
3. Which environments follow similar mutant orderings, and how does accounting
   for measurement uncertainty affect those groups and trajectories?

## Load package and data

```{r load}
if (requireNamespace("MPCurver", quietly = TRUE)) {
  library(MPCurver)
} else if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(".", quiet = TRUE)
} else {
  stop("Need either an installed MPCurver package or devtools::load_all().")
}

set.seed(42)

locate_elife_csv <- function() {
  pkg_path <- system.file("extdata", "elife-61271-fig2-data1-v2.csv", package = "MPCurver")
  if (nzchar(pkg_path) && file.exists(pkg_path)) {
    return(pkg_path)
  }

  src_path <- file.path("inst", "extdata", "elife-61271-fig2-data1-v2.csv")
  if (file.exists(src_path)) {
    return(src_path)
  }

  stop("Could not locate elife-61271-fig2-data1-v2.csv.")
}

csv_path <- locate_elife_csv()
dat <- read.csv(csv_path, check.names = FALSE)

fitness_cols <- grep("_fitness$", names(dat), value = TRUE)
error_cols <- grep("_error$", names(dat), value = TRUE)

X <- as.matrix(dat[, fitness_cols, drop = FALSE])
mode(X) <- "numeric"
colnames(X) <- sub("_fitness$", "", fitness_cols)

S <- as.matrix(dat[, error_cols, drop = FALSE])
mode(S) <- "numeric"
colnames(S) <- sub("_error$", "", error_cols)

meta <- dat[, c("barcode", "gene", "type", "ploidy", "class", "mutation_type"), drop = FALSE]
```


For this illustration, we calibrate the reported errors using a Student's
$t$ distribution with four degrees of freedom. For each ratio $z_t=X/S$, we
find the Gaussian score $z_n=\Phi^{-1}\{F_{t_4}(z_t)\}$ with the same cumulative
probability and use the adjusted scale $\widetilde S=S z_t/z_n$. The choice of
four degrees of freedom is an analysis assumption; the resulting scales are
treated as fixed measurement standard deviations in the fits below.

```{r adjust-se}
t_adjust_se <- function(Y, S, nu, tol = 1e-12) {
  if (!is.matrix(Y) || !is.numeric(Y)) {
    stop("`Y` must be a numeric matrix.")
  }
  if (!is.matrix(S) || !is.numeric(S)) {
    stop("`S` must be a numeric matrix.")
  }
  if (!all(dim(Y) == dim(S))) {
    stop("`Y` and `S` must have the same dimensions.")
  }
  if (any(S < 0, na.rm = TRUE)) {
    stop("`S` must be nonnegative.")
  }
  
  if (length(nu) == 1L) {
    if (!is.numeric(nu) || is.na(nu) || nu <= 0) {
      stop("`nu` must be a positive number.")
    }
    nu_mat <- matrix(nu, nrow = nrow(Y), ncol = ncol(Y))
  } else {
    if (!is.matrix(nu) || !is.numeric(nu) || !all(dim(nu) == dim(Y))) {
      stop("`nu` must be either a positive scalar or a numeric matrix with the same dimensions as `Y`.")
    }
    if (any(nu <= 0, na.rm = TRUE)) {
      stop("All entries of `nu` must be positive.")
    }
    nu_mat <- nu
  }
  
  z_t <- Y / S
  
  # Main calculation:
  # S_tilde = S * z_t / qnorm(pt(z_t, df = nu))
  p_t <- pt(z_t, df = nu_mat)
  z_n <- qnorm(p_t)
  
  S_tilde <- S * z_t / z_n
  
  # Numerical fix for z_t close to 0:
  # limit of z / qnorm(pt(z, df=nu)) as z -> 0 is dnorm(0) / dt(0, df=nu)
  near_zero <- is.finite(z_t) & abs(z_t) < tol
  if (any(near_zero, na.rm = TRUE)) {
    limit_factor <- dnorm(0) / dt(0, df = nu_mat[near_zero])
    S_tilde[near_zero] <- S[near_zero] * limit_factor
  }
  
  # Preserve zeros
  zero_se <- (S == 0) & !is.na(S)
  if (any(zero_se)) {
    S_tilde[zero_se] <- 0
  }
  
  dimnames(S_tilde) <- dimnames(S)
  S_tilde
}
S <- t_adjust_se(X, S, nu = 4)
```



In this dataset:

- each row is a mutant;
- each column is an environment;
- each matrix entry is a fitness score;
- and the paired `*_error` columns provide observation-level measurement
  error estimates used to construct the adjusted standard deviations `S`.

## First look at the data

```{r helpers, include=FALSE}
environment_category <- function(env) {
  if (grepl("Ferm|Day", env)) return("time course")
  if (grepl("NaCl|KCl", env)) return("salt stress")
  if (grepl("Ben|Flu|GdA|DMSO", env)) return("drug / chemical")
  if (grepl("Gluc|Baffle", env)) return("glucose / batch")
  if (grepl("Raf|Gly|EtOH|Suc", env)) return("carbon source")
  "other"
}

env_category <- vapply(colnames(X), environment_category, character(1))

plot_ordered_heatmap <- function(X, order_idx, env_idx = seq_len(ncol(X)),
                                 main = "", zlim = c(-2.5, 2.5)) {
  Xsub <- X[order_idx, env_idx, drop = FALSE]
  Xsub <- scale(Xsub)
  Xsub[!is.finite(Xsub)] <- 0
  pal <- colorRampPalette(c("#313695", "#74add1", "#f7f7f7", "#f46d43", "#a50026"))(101)
  image(
    x = seq_len(nrow(Xsub) + 1L),
    y = seq_len(ncol(Xsub) + 1L),
    z = Xsub,
    col = pal,
    axes = FALSE,
    xlab = "Mutants ordered by inferred pseudotime",
    ylab = "",
    zlim = zlim,
    main = main
  )
  axis(2, at = seq_len(ncol(Xsub)), labels = colnames(Xsub), las = 2, cex.axis = 0.55)
  box()
}

```

```{r data-summary}
data.frame(
  quantity = c("mutants", "environments", "fitness columns", "error columns"),
  value = c(nrow(X), ncol(X), length(fitness_cols), length(error_cols))
) |>
  knitr::kable()

sort(table(env_category), decreasing = TRUE) |>
  as.data.frame() |>
  stats::setNames(c("category", "n_environments")) |>
  knitr::kable()
```

This is a good MPCurver example because the environments span several broad
condition types, while the mutants are shared across all columns. That makes it
natural to ask whether the same latent mutant ordering explains every
environment, or whether different subsets of environments prefer different
orderings.

## Step 1: Fit a single-ordering MPCurve

We start with the simplest question: can one latent ordering explain the main
structure in the dataset?

The fitting choices are:

- `method = "isomap"` supplies the initialization;
- `K = 50` uses a 50-bin latent grid;
- `rw_q = 2` uses a second-order random-walk prior for smooth trajectories.

```{r uni-fit}
fit_unidir <- fit_mpcurve(
  X,
  method = "isomap",
  K = 50,
  rw_q = 2,
  verbose = FALSE
)
```

### Check convergence

The ELBO trace shows progress in fitting the variational model. A plateau
indicates small improvements between successive iterations.

```{r uni-convergence, out.width = "100%", fig.alt = "ELBO trace for the single-ordering CAVI fit, showing convergence over iterations."}
plot(fit_unidir, plot_type = "elbo")
```

### Inspect a few fitted trajectories

```{r uni-trajectories, fig.height = 8, out.width = "100%", fig.alt = "Four panels showing fitted single-ordering CAVI trajectories for the first four environment dimensions."}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_unidir, dims = 1)
plot(fit_unidir, dims = 2)
plot(fit_unidir, dims = 3)
plot(fit_unidir, dims = 4)
par(mfrow = c(1, 1))
```

The first fit already recovers coherent structure, but the trajectories are
still rougher than we would like for interpretation. 
This is a common pattern on real data: the latent ordering may be sensible even when the fitted curves need stronger regularization.

## Step 2: Encourage smoother trajectories

We now refit the same single-ordering model, but add a stronger prior on
`lambda_j` through `lambda_sd_prior_rate`. Larger values encourage smoother
trajectories.

```{r uni-fit-smooth}
fit_unidir_smooth <- fit_mpcurve(
  X,
  method = "isomap",
  K = 50,
  iter = 500,
  rw_q = 2,
  lambda_sd_prior_rate = 1e4,
  verbose = FALSE
)
```

```{r uni-convergence-smooth, out.width = "100%", fig.alt = "ELBO trace for the smoothed single-ordering CAVI fit."}
plot(fit_unidir_smooth, plot_type = "elbo")
fit_unidir_smooth
```

```{r uni-trajectories-smooth, fig.height = 8, out.width = "100%", fig.alt = "Four panels showing fitted trajectories under the smoothed single-ordering CAVI fit."}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_unidir_smooth, dims = 1)
plot(fit_unidir_smooth, dims = 3)
plot(fit_unidir_smooth, dims = 7)
plot(fit_unidir_smooth, dims = 9)
par(mfrow = c(1, 1))
```

These curves are easier to read: the global ordering remains, but the fitted
environment trajectories are less wiggly.

A useful comparison is whether smoothing changes the inferred mutant ordering
itself, or mainly changes the trajectory estimates.

```{r compare-single-orderings, out.width = "100%", fig.alt = "Scatter plot comparing inferred mutant pseudotime between the default and smoothed single-ordering fits."}
cor(
  fit_unidir_smooth$locations$mean$pseudotime,
  fit_unidir$locations$mean$pseudotime
)

plot(
  fit_unidir_smooth$locations$mean$pseudotime,
  fit_unidir$locations$mean$pseudotime,
  xlab = "Latent position (smoother fit)",
  ylab = "Latent position (less regularized fit)"
)
abline(0, 1, lty = 2)
```

The correlation and scatterplot show how closely the two mutant orderings
agree. Together with the trajectory plots, they help distinguish a change
in curve smoothness from a change in sample ordering.

## Step 3: Explore two orderings and environment groups

The single-ordering fit assumes that every environment varies smoothly along
the same latent mutant axis. A two-ordering model lets groups of environments
follow different axes. We specify `intrinsic_dim = 2` and jointly estimate
two mutant orderings, environment-assignment probabilities, and trajectories
conditional on each assignment.

Structural variational inference keeps the dependence between an environment's
assignment and its trajectory. Each environment has one estimated noise
variance shared across orderings, while its smoothness can differ between
orderings. This analysis explores a two-axis representation; choosing the
number of axes would require a separate model-selection assessment.

```{r partition-fit}
fit_partition <- fit_mpcurve(
  X,
  method = "isomap",
  intrinsic_dim = 2,
  ridge = 0,
  K = 50,
  iter = 500,
  rw_q = 2,
  lambda_sd_prior_rate = 1e4,
  verbose = FALSE
)
```

```{r partition-summary}
fit_partition
```

### Inspect trajectories under each ordering

For `plot_type = "mu"`, each panel shows the conditional posterior mean
$E[U_j\mid Z_j=m]$: the estimated environment trajectory if it follows
ordering $m$. The title reports the assignment probability $q(Z_j=m)$,
indicating the support for that interpretation.

```{r partition-plot, out.width = "100%", fig.alt = "Conditional posterior mean trajectory for environment dimension 1 under each of two fixed orderings."}
plot(fit_partition, plot_type = "mu", dims = 1)
```

### Which environments were assigned to which ordering?

The table lists each environment's most probable ordering. To examine
uncertain assignments, inspect the corresponding row of
`fit_partition$partition$pi_weights`.

```{r partition-table}
assignment_tbl <- data.frame(
  environment = colnames(X),
  category = env_category,
  ordering = fit_partition$partition$assign,
  stringsAsFactors = FALSE
)

assignment_tbl[order(assignment_tbl$ordering, assignment_tbl$category, assignment_tbl$environment), ] |>
  knitr::kable(row.names = FALSE)
```

### How similar are the inferred orderings?

Compare the inferred mutant positions across orderings. A correlation near
$1$ or $-1$ indicates similar positions after allowing for reversal of the
axis; smaller absolute correlations indicate more distinct orderings.

```{r partition-correlation}
ordering_names <- names(fit_partition$locations)
ordering_names <- ordering_names[ordering_names %in% unique(fit_partition$partition$assign)]

t_df <- setNames(
  as.data.frame(lapply(ordering_names, function(nm) {
    fit_partition$locations[[nm]]$mean$pseudotime
  })),
  ordering_names
)

cor(t_df) |>
  knitr::kable(digits = 3)
```



## Step 4: Account for measurement uncertainty

We now supply the adjusted measurement standard deviations computed above
through `S`. This gives more precise observations greater influence in the
likelihood. Comparing these fits with the estimated-noise fits shows how
the supplied measurement precision affects orderings and trajectories.

### Single-ordering fit with known measurement SD

```{r uni-fit-known-sd}
fit_unidir_known_sd <- fit_mpcurve(
  X,
  S = S,
  method = "isomap",
  K = 50,
  iter = 500,
  rw_q = 2,
  lambda_sd_prior_rate = 1e4,
  verbose = FALSE
)
```

```{r uni-known-sd-convergence, out.width = "100%", fig.alt = "ELBO trace for the single-ordering CAVI fit with known measurement standard deviations."}
plot(fit_unidir_known_sd, plot_type = "elbo")
```

```{r uni-known-sd-trajectories, fig.height = 8, out.width = "100%", fig.alt = "Four panels showing fitted trajectories for the single-ordering CAVI fit with known measurement SDs."}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(fit_unidir_known_sd, dims = 1)
plot(fit_unidir_known_sd, dims = 2)
plot(fit_unidir_known_sd, dims = 31)
plot(fit_unidir_known_sd, dims = 45)
par(mfrow = c(1, 1))
```

### Two-ordering fit with supplied measurement SD

We refit the two-ordering model with the same smoothing settings and the
supplied standard deviations. Examine both the trajectories and assignment
probabilities to see how the environment groups respond to this noise model.

```{r partition-fit-known-sd}
fit_partition_known_sd <- fit_mpcurve(
  X,
  S = S,
  method = "isomap",
  intrinsic_dim = 2,
  ridge = 0,
  K = 50,
  iter = 500,
  rw_q = 2,
  lambda_sd_prior_rate = 1e4,
  verbose = FALSE
)

fit_partition_known_sd
```

```{r known-sd-partition-trajectories, fig.height = 5.5, out.width = "100%", fig.alt = "Conditional posterior mean trajectory for environment dimension 31 under each of two fixed orderings with known measurement SDs."}
plot(fit_partition_known_sd, plot_type = "mu", dims = 31)
```



## Takeaways

This dataset gives a useful template for real-world `fit_mpcurve()` analyses.

1. Start with a single-ordering fit to establish the basic latent ordering.
2. Use `lambda_sd_prior_rate` when the ordering looks sensible but the fitted
   trajectories are too rough to interpret comfortably.
3. Use `intrinsic_dim` to explore multiple orderings. Interpret the environment
   groups together with their assignment probabilities and fitted trajectories.
4. When measurement standard deviations are available, supply them through
   `S` and examine how the inferred structure changes. Document any calibration
   assumptions used to obtain those standard deviations.
