---
title: "Estimating Intrinsic Dimension with MPCurver"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Estimating Intrinsic Dimension with MPCurver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 7, fig.height = 5)
```

When the intrinsic dimension is unknown, start with
`fit_mpcurve(X, intrinsic_dim = "auto")` and its default adaptive partition
prior. This is computationally simpler than fitting a sequence of candidate
dimensions: feature similarity chooses an initial dimension, then one adaptive
fit estimates ordering weights and removes numerical-zero orderings.

Use `select_mpcurve_dimension()` when your analysis calls for explicit model
comparison under a fixed uniform feature-assignment prior. The two approaches
use different priors and selection rules and can produce different dimensions.
Supplying an integer `intrinsic_dim = M` instead fits that specified dimension.
This tutorial shows both estimation approaches;
[Multiple orderings](partition.html) introduces the model and software use.
The [algorithm details](cavi_partition.html) are available separately.

## Example data

We simulate 200 samples and twelve features, with two groups following
different latent orderings. The generating dimension is two; it is not
supplied to either fitting procedure below.

```{r data}
library(MPCurver)
sim <- simulate_dual_trajectory(
  n = 200,
  d1 = 6,
  d2 = 6,
  d_noise = 0,
  noise_sd = 0.05,
  trajectory_family = c("quadratic", "monotone"),
  seed = 1
)
X <- sim$X
```

## Automatic initialization and adaptive fitting

Use `intrinsic_dim = "auto"` to choose the starting model dimension from the
feature-similarity tree. `max_intrinsic_dim` bounds the candidate dimensions;
here we consider at most four orderings and use twenty position bins.
The default adaptive feature-assignment prior is used during fitting.

```{r automatic-fit}
fit_auto <- fit_mpcurve(
  X,
  intrinsic_dim = "auto",
  num_bins = 20,
  max_iter = 200,
  init_control = mpcurve_init_control(max_intrinsic_dim = 4)
)
```

Automatic initialization uses spline-R-squared feature similarity by default.
It evaluates hierarchical-clustering cuts and selects the eligible cut with
the largest mean silhouette. Every cluster must contain at least
`init_control$min_cluster_size` features (default 2). If no multi-ordering cut
satisfies this rule, it fits one ordering.

For cluster sizes $P_1,\ldots,P_M$, the adaptive fit initializes the ordering
prior weights as $\omega_m=P_m/P$, where $P$ is the total number of features.
Every feature starts from this same probability vector. After fitting,
orderings with numerical-zero estimated prior weights are removed. The returned
model contains the estimated number of orderings.

The minimum cluster size prevents a singleton feature from serving as the
sole evidence for a latent ordering: sorting that feature by its own values
would produce a smooth trajectory by construction. A singleton can reflect
noise, weak measurement, or model mismatch; MPCurver has no separate
background-feature component.

## Inspect the estimated model {#report-the-supported-dimension}

`fit_auto$intrinsic_dim` is the number of orderings in the returned model.
Table 1 reports this estimated dimension.

```{r dimensions, echo=FALSE}
knitr::kable(
  data.frame(
    Quantity = "Estimated intrinsic dimension",
    Field = "intrinsic_dim",
    Value = fit_auto$intrinsic_dim
  ),
  caption = "Table 1. Estimated intrinsic dimension of the returned model."
)
```

For ordering $m$, empirical Bayes estimates its prior weight as

$$
\widehat\omega_m=\frac{1}{P}\sum_{j=1}^{P} q(Z_j=m),
$$

where $P$ is the number of features. Automatic adaptive fitting removes an
ordering when $\widehat\omega_m$ is at or below
`control$effective_count_tol / P`. The tolerance defaults to `1e-8` and
identifies estimated prior weights that are numerically close to zero.
Automatic adaptive pruning warns when the tolerance exceeds `1e-6`, before
division by $P$, because large values may remove supported orderings.

Empty orderings are removed without further fitting. The remaining assignment
probabilities and prior weights are normalized, and the final objective is
recomputed. Initial dimensions and any removed orderings are recorded in
`fit_auto$dimension_estimation`.

Table 2 shows the estimated prior weights and their corresponding expected
feature counts, $P\widehat\omega_m$.

```{r ordering-support}
feature_counts <- colSums(fitted_assignments(fit_auto))
# At the adaptive update, the estimated prior weight equals the mean
# posterior assignment probability; this also covers a single retained ordering.
prior_weights <- colMeans(fitted_assignments(fit_auto))
```

```{r support-table, echo=FALSE}
knitr::kable(
  data.frame(Ordering = names(feature_counts),
             Prior_weight = as.numeric(prior_weights),
             Expected_feature_count = as.numeric(feature_counts)),
  digits = 6,
  caption = "Table 2. Estimated ordering prior weights and corresponding expected feature counts."
)
```

Use `fitted_positions()`, `fitted_trajectories()`, and `fitted_assignments()`
to extract results with consistent dimensions even when the estimated dimension
is one.
Supplying an integer `intrinsic_dim` retains that specified dimension.
With a fixed feature-assignment prior, automatic fitting retains the
dimension chosen at initialization.

## Compare dimensions under a uniform prior

`select_mpcurve_dimension()` fits adjacent model dimensions with a fixed
uniform feature-assignment prior and compares final soft-assignment objectives
at $T=1$. Forward selection starts at $M=1$ and accepts the next dimension
only when its objective is strictly larger. Backward selection starts at
`max_intrinsic_dim` and tests smaller dimensions by the same rule.

```{r select-dimension}
selected <- select_mpcurve_dimension(
  X,
  max_intrinsic_dim = 4,
  direction = "forward",
  num_bins = 20,
  max_iter = 200
)
selected$intrinsic_dim
```

Each candidate $M$ is fitted once. For $M > 1$, features are grouped by
similarity, and the selected `initial_method` initializes each group.
All candidates use a common grid size and the supplied model and fitting
settings. With uniform prior weights,
an ordering receiving no feature mass still changes the assignment-prior
term by $P\log\{M/(M-1)\}$ relative to the smaller model.

Table 3 records the objective comparisons, including the rejected stopping
step when present. Table 4 reports each attempted fit and its convergence
status.

```{r inspect-history, eval=FALSE}
selected$dimension_selection$history
```

<div class="table-responsive">

```{r selection-history, echo=FALSE}
knitr::kable(
  selected$dimension_selection$history[, c("current_M", "candidate_M",
    "current_score", "candidate_score", "delta", "accepted")],
  col.names = c("From M", "To M", "Current objective", "Candidate objective",
                "Change", "Accepted"),
  digits = 3,
  caption = "Table 3. Adjacent-dimension objective comparisons in forward selection."
)
```

</div>

```{r inspect-candidates, eval=FALSE}
selected$dimension_selection$candidates
```

<div class="table-responsive">

```{r selection-candidates, echo=FALSE}
candidates <- selected$dimension_selection$candidates
knitr::kable(
  candidates[, c("M", "initialization", "status", "score", "converged")],
  col.names = c("M", "Initialization", "Status", "Objective", "Converged"),
  digits = 3,
  caption = "Table 4. Initialization, fitting objective, and convergence status for each candidate dimension."
)
```

</div>

The complete candidate record also contains warnings and errors. Greedy
selection depends on initialization and convergence and can stop at a local
choice. Inspect these records and, for candidates that reach their iteration
limit, increase `max_iter` before relying on the comparison. The adaptive
occupancy count and uniform-prior search use different criteria.
