---
title: "Getting Started with MPCurver"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with MPCurver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

```{r load}
library(MPCurver)
set.seed(42)
```

`MPCurver` estimates an ordering of samples and smooth feature trajectories
along that ordering. This tutorial introduces the model, then uses a simulated
spiral to show how to fit a model, assess convergence, and extract results.
The [algorithm details](cavi_single.html) are available separately.

## Statistical model

We observe a data matrix
\[
X = (x_{ij}) \in \mathbb{R}^{n \times d},
\]
where row \(i\) is one sample (for example, a cell) and column \(j\) is one feature.

The single-ordering MPCurve model assumes:

1. each cell has a latent position
   \[
   c_i \in \{1,\dots,K\},
   \]
2. each feature has a smooth trajectory over the \(K\) ordered components,
   written as
   \[
   \boldsymbol{u}_j = (u_{j1},\dots,u_{jK})^\top \in \mathbb{R}^K,
   \]
3. observations follow a Gaussian likelihood
   \[
   x_{ij} \mid c_i = k, \boldsymbol{u}_j, \sigma_j^2 \sim N(u_{jk}, \sigma_j^2),
   \]
4. and each trajectory is regularized by a Gaussian Markov random field prior
   along the component axis:
   \[
   p(\boldsymbol{u}_j\mid\lambda_j)
   \propto \exp\!\left(-\frac{\lambda_j}{2}
   \boldsymbol{u}_j^\top Q_K\boldsymbol{u}_j\right),
   \]
   where \(Q_K\) is the random-walk precision matrix on the \(K\) bins.
   The default intrinsic prior penalizes differences along the trajectory;
   the likelihood identifies directions left unpenalized by the prior.

The model estimates a separate noise variance \(\sigma_j^2\) and smoothness
parameter \(\lambda_j\) for each feature. The noise variance is shared across
grid positions. The position prior is $\Pr(c_i=k)=\pi_k$, with $\sum_k\pi_k=1$.
By default these probabilities are also estimated from the data.

The fitting algorithm estimates position and trajectory distributions together
with the model parameters. Its objective, coordinate updates, and stopping
rule are described in [Single-ordering CAVI](cavi_single.html).

## Simulated spiral example

We simulate a two-dimensional spiral with:

- `n = 1000` cells,
- latent range `t in [1.5*pi, 4.5*pi]`,
- measurement standard deviation `sigma = 0.12`,
- and a fitting grid of `num_bins = 60` positions.

```{r simulate}
sim <- simulate_swiss_roll_1d_2d(
  n = 1000,
  t_range = c(1.5 * pi, 4.5 * pi),
  noise_sd = 0.12,
  seed = 1
)

X <- as.matrix(sim$obs)
t_true <- sim$t

cat("Data dimensions:", nrow(X), "x", ncol(X), "\n")
cat("t range:", round(range(t_true), 3), "\n")
```

Figure 1 shows simulated spiral observations, colored by the generating latent position.

```{r plot-data, fig.cap="Figure 1. Simulated spiral observations, colored by the generating latent position."}
pal <- grDevices::colorRampPalette(
  c("#1F3A5F", "#2E8A99", "#F2C14E", "#E76F51")
)(256)
col_true <- pal[cut(t_true, breaks = 256, labels = FALSE)]

plot(X, col = col_true, pch = 19, cex = 0.55,
     xlab = "x", ylab = "y",
     main = "Observed spiral, coloured by true pseudotime")
```

## Fit the model

For your own data, supply a numeric matrix or data frame with samples in rows
and features in columns. Values must be finite: handle missing values before
fitting. MPCurver does not impute or standardize the fitting data. Optional
measurement standard deviations `S` use the same units and positional order
as `X`, as a feature vector or a matrix of the same dimensions. See
`?fit_mpcurve` for the complete input contract.

```{r fit-cavi}
fit <- fit_mpcurve(
    X,
    initial_method = "fiedler",
    num_bins = 60,
    max_iter = 120
  )

stopifnot(inherits(fit, "mpcurve"))
stopifnot(identical(fit$algorithm, "cavi"))

fit
```

```{r summary-fit}
summary(fit)
```

## Check convergence and recovery

First, check whether the variational objective increases throughout fitting.

```{r check-elbo}
stopifnot(all(diff(fit$elbo_trace) >= -1e-8))
cat("ELBO monotone:", TRUE, "\n")
cat("ELBO range:", round(range(fit$elbo_trace), 3), "\n")
```

We can also compare the inferred ordering to the true latent parameter using
the posterior-mean pseudotime on the normalized grid.

```{r ordering-recovery}
pseudotime_hat <- fitted_positions(fit)[, 1]
rho_t <- abs(cor(pseudotime_hat, t_true, method = "spearman"))

cat("Absolute Spearman correlation with true t:", round(rho_t, 4), "\n")
```

## Visualize the fitted ordering

Figure 2 shows true pseudotime (left) and inferred CAVI pseudotime (right).

```{r plot-fit, fig.cap="Figure 2. True pseudotime (left) and inferred CAVI pseudotime (right)."}
col_fit <- pal[cut(pseudotime_hat, breaks = 256, labels = FALSE)]

op <- par(mfrow = c(1, 2), mar = c(4, 4, 3.5, 1))
plot(X, col = col_true, pch = 19, cex = 0.5,
     xlab = "x", ylab = "y", main = "Truth")
plot(X, col = col_fit, pch = 19, cex = 0.5,
     xlab = "x", ylab = "y", main = "CAVI fit")
par(op)
```

The fitted object also provides plots of the objective and posterior mean
trajectory.

Figure 3 shows the CAVI trace plot from the fitted mpcurve object.

```{r plot-trace, fig.cap="Figure 3. CAVI trace plot from the fitted mpcurve object."}
plot(fit, plot_type = "elbo")
```

Figure 4 shows posterior mean trajectories across the K bins.

```{r plot-mu, fig.cap="Figure 4. Posterior mean trajectories across the K bins."}
plot(fit, plot_type = "mu")
```

## Using the results

The ordering-recovery statistic above compares estimated sample positions
with the generating latent variable. Its absolute value accounts for the
arbitrary direction of an inferred ordering. In an application, use known
sample annotations to help interpret that direction and assess whether
nearby samples have similar feature profiles.

The extraction functions use the same dimensions for one or multiple orderings:

```{r extract-results}
positions <- fitted_positions(fit)                 # samples x orderings
trajectories <- fitted_trajectories(fit)            # features x bins x orderings
trajectory_sd <- fitted_trajectories(fit, "sd")
assignment_probabilities <- fitted_assignments(fit) # features x orderings
head(positions)
```

Use `fitted_positions(fit, "probability")` for the full position probabilities
and `fitted_trajectories(fit, "covariance")` for trajectory covariance matrices.
Input sample and feature names are preserved. Even a single ordering retains
its matrix or array dimension. If the objective is still increasing at the
iteration limit, continue with `fit <- do_mpcurve(fit, max_iter = 100)`.
Continuation inherits the fitted settings. Use
`control = mpcurve_continue_control(lambda_init = 5, fix_lambda = TRUE)`
to reset and fix precision at 5, or `fix_lambda = FALSE` to resume its updates.
Changing the random-walk order or measurement errors requires a new fit.

When different feature groups may follow different sample orderings, see the
[multiple-ordering tutorial](partition.html). To choose the dimension, see
[Estimating intrinsic dimension](intrinsic_dimension.html). For data with supplied
measurement uncertainty, see the [fitness analysis](fitness.html).
