---
title: "Getting Started with MPCurver"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with MPCurver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

```{r load}
library(MPCurver)
set.seed(42)
```

`MPCurver` estimates an ordering of samples and smooth feature trajectories
along that ordering. This vignette introduces the statistical model and its
coordinate-ascent variational inference (CAVI) algorithm, then uses a simulated
spiral to show how to fit a model, assess convergence, and interpret pseudotime.

## 1  Statistical model

We observe a data matrix
\[
X = (x_{ij}) \in \mathbb{R}^{n \times d},
\]
where row \(i\) is one sample (for example, a cell) and column \(j\) is one feature.

The single-ordering MPCurve model assumes:

1. each cell has a latent position
   \[
   c_i \in \{1,\dots,K\},
   \]
2. each feature has a smooth trajectory over the \(K\) ordered components,
   written as
   \[
   \boldsymbol{u}_j = (u_{j1},\dots,u_{jK})^\top \in \mathbb{R}^K,
   \]
3. observations follow a Gaussian likelihood
   \[
   x_{ij} \mid c_i = k, \boldsymbol{u}_j, \sigma_j^2 \sim N(u_{jk}, \sigma_j^2),
   \]
4. and each trajectory is regularized by a Gaussian Markov random field prior
   along the component axis:
   \[
   p(\boldsymbol{u}_j\mid\lambda_j)
   \propto \exp\!\left(-\frac{\lambda_j}{2}
   \boldsymbol{u}_j^\top Q_K\boldsymbol{u}_j\right),
   \]
   where \(Q_K\) is the random-walk precision matrix on the \(K\) bins.
   The default intrinsic prior penalizes differences along the trajectory;
   the likelihood identifies directions left unpenalized by the prior.

The model estimates a separate noise variance \(\sigma_j^2\) and smoothness
parameter \(\lambda_j\) for each feature. The noise variance is shared across
grid positions. To describe uncertainty in positions and trajectories, CAVI
uses the variational family

\[
q(C,U) = \prod_{i=1}^n q_i(c_i)\prod_{j=1}^d q_j(\boldsymbol{u}_j).
\]

## 2  Variational algorithm

`fit_mpcurve()` alternates between updates of the posterior distributions and
model parameters to maximize the evidence lower bound (ELBO). Write
\[
r_{ik} := q_i(c_i = k),
\qquad
N_k := \sum_{i=1}^n r_{ik},
\]
and for each feature \(j\),
\[
q_j(\boldsymbol{u}_j) = N(\boldsymbol{m}_j, S_j).
\]

Under the mean-field family
\[
q(C,U)=\prod_{i=1}^n q_i(c_i)\prod_{j=1}^d q_j(\boldsymbol{u}_j),
\]
the ELBO can be written, up to additive constants, as
\[
\mathcal L(q)
 =
\sum_{i=1}^n\sum_{k=1}^K r_{ik}\log \pi_k
- \sum_{i=1}^n\sum_{k=1}^K r_{ik}\log r_{ik}
\]
\[
- \frac12 \sum_{i=1}^n\sum_{k=1}^K\sum_{j=1}^d
r_{ik}
\left[
\log(2\pi \sigma_j^2)
+
\frac{(x_{ij}-m_{jk})^2 + S_{j,kk}}{\sigma_j^2}
\right]
\]
\[
+ \frac12 \sum_{j=1}^d
\left[
\operatorname{rank}(Q_K)\log \lambda_j
- \lambda_j\Big(\boldsymbol{m}_j^\top Q_K \boldsymbol{m}_j + \operatorname{tr}(Q_K S_j)\Big)
+ \log |S_j|
\right].
\]

This objective, including its additive constants, is recorded in `elbo_trace`.
For the intrinsic prior, normalization uses the rank and generalized
determinant of \(Q_K\), with a fixed reference measure on its null space.
It provides a within-model fitting objective. The updates below maximize it
one block at a time, with estimated noise, adaptive position probabilities,
and no additional smoothness-parameter prior. Parameter estimates are
restricted to the configured bounds.

### 2.1  Update for \(q(\boldsymbol{u}_j)\)

Fix \(r_{ik}\), \(\sigma_j^2\), and \(\lambda_j\). The ELBO terms involving
\(\boldsymbol{u}_j\) are quadratic, so the optimal variational factor is Gaussian:
\[
\log q_j^\star(\boldsymbol{u}_j)
=
-\frac12 \boldsymbol{u}_j^\top A_j \boldsymbol{u}_j
+ \boldsymbol{u}_j^\top \boldsymbol{b}_j + \text{const},
\]
where
\[
A_j
=
\lambda_j Q_K +
\operatorname{diag}\!\left(N_1/\sigma_j^2,\dots,N_K/\sigma_j^2\right),
\]
and
\[
b_{jk}
=
\frac{1}{\sigma_j^2}\sum_{i=1}^n r_{ik} x_{ij}.
\]
Therefore,
\[
S_j = A_j^{-1},
\qquad
\boldsymbol{m}_j = S_j \boldsymbol{b}_j.
\]

The diagonal likelihood term supplies information at the grid positions,
while the random-walk precision shares information between neighboring
positions. Their combination determines both the fitted trajectory and its
uncertainty.

### 2.2  Update for \(q(c_i)\)

Fix \(q(U)\), \(\pi\), and \(\sigma_j^2\). Collecting the ELBO terms involving
the row \(r_{i\cdot}\) gives
\[
\log r_{ik}
= a_i +
\log \pi_k
- \frac12 \sum_{j=1}^d
\left[
\log(2\pi \sigma_j^2)
+
\frac{(x_{ij}-m_{jk})^2 + S_{j,kk}}{\sigma_j^2}
\right].
\]
So the cell-position posterior is updated by a row-wise softmax:
\[
r_{i\cdot} \leftarrow \operatorname{softmax}\!\big(\ell_{i1},\dots,\ell_{iK}\big),
\]
with \(\ell_{ik}\) equal to the position-specific terms above and \(a_i\)
the normalizing constant.

The \(S_{j,kk}\) term accounts for trajectory uncertainty when assigning a
sample to a grid position.

### 2.3  Update for \(\pi_k\)

Fixing everything else, we maximize
\[
\sum_{k=1}^K N_k \log \pi_k
\]
subject to \(\sum_k \pi_k = 1\). A Lagrange multiplier gives
\[
\pi_k = \frac{N_k}{n}.
\]

### 2.4  Update for \(\sigma_j^2\)

Fixing \(q(U)\) and \(q(C)\), differentiate the ELBO with respect to
\(\sigma_j^2\). The stationary point is
\[
\sigma_j^2
=
\frac{1}{n}
\sum_{i=1}^n \sum_{k=1}^K
r_{ik}\Big[(x_{ij}-m_{jk})^2 + S_{j,kk}\Big].
\]

So each feature variance is updated from the posterior expected residual sum of
squares.

### 2.5  Update for \(\lambda_j\)

Fixing \(q(U)\), only the prior part of the ELBO depends on \(\lambda_j\):
\[
\frac12 \operatorname{rank}(Q_K)\log \lambda_j
- \frac12 \lambda_j
\Big(\boldsymbol{m}_j^\top Q_K \boldsymbol{m}_j + \operatorname{tr}(Q_K S_j)\Big).
\]
Setting the derivative to zero yields
\[
\lambda_j
=
\frac{\operatorname{rank}(Q_K)}
{\boldsymbol{m}_j^\top Q_K \boldsymbol{m}_j + \operatorname{tr}(Q_K S_j)}.
\]

### 2.6  Assessing convergence

Each update maximizes one objective block while holding the others fixed, so
`elbo_trace` should be nondecreasing up to numerical tolerance. A plateau
indicates that successive iterations are making small improvements. Because
the objective can have multiple optima, compare several initializations when
the trajectory is complex or weakly identified.

## 3  Simulated spiral example

We simulate a two-dimensional spiral with:

- `n = 1000` cells,
- latent range `t in [1.5*pi, 4.5*pi]`,
- measurement standard deviation `sigma = 0.12`,
- and a fitting grid of `K = 60` positions.

```{r simulate}
sim <- simulate_swiss_roll_1d_2d(
  n = 1000,
  t_range = c(1.5 * pi, 4.5 * pi),
  sigma = 0.12,
  seed = 1
)

X <- as.matrix(sim$obs)
t_true <- sim$t

cat("Data dimensions:", nrow(X), "x", ncol(X), "\n")
cat("t range:", round(range(t_true), 3), "\n")
```

```{r plot-data, fig.cap="Simulated spiral observations, colored by the generating latent position."}
pal <- grDevices::colorRampPalette(
  c("#1F3A5F", "#2E8A99", "#F2C14E", "#E76F51")
)(256)
col_true <- pal[cut(t_true, breaks = 256, labels = FALSE)]

plot(X, col = col_true, pch = 19, cex = 0.55,
     xlab = "x", ylab = "y",
     main = "Observed spiral, coloured by true pseudotime")
```

## 4  Fit the model

```{r fit-cavi}
fit <- fit_mpcurve(
  X,
  method = "fiedler",
  K = 60,
  iter = 120
)

stopifnot(inherits(fit, "mpcurve"))
stopifnot(identical(fit$algorithm, "cavi"))

fit
```

```{r summary-fit}
summary(fit)
```

## 5  Check convergence and recovery

First, check whether the variational objective increases throughout fitting.

```{r check-elbo}
stopifnot(all(diff(fit$elbo_trace) >= -1e-8))
cat("ELBO monotone:", TRUE, "\n")
cat("ELBO range:", round(range(fit$elbo_trace), 3), "\n")
```

We can also compare the inferred ordering to the true latent parameter using
the posterior-mean component index
\[
\hat c_i = \sum_{k=1}^K k \, r_{ik}.
\]

```{r ordering-recovery}
pseudotime_hat <- as.numeric(fit$gamma %*% seq_len(ncol(fit$gamma)))
rho_t <- abs(cor(pseudotime_hat, t_true, method = "spearman"))

cat("Absolute Spearman correlation with true t:", round(rho_t, 4), "\n")
```

## 6  Visualize the fitted ordering

```{r plot-fit, fig.cap="True pseudotime (left) and inferred CAVI pseudotime (right)."}
col_fit <- pal[cut(pseudotime_hat, breaks = 256, labels = FALSE)]

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
plot(X, col = col_true, pch = 19, cex = 0.5,
     xlab = "x", ylab = "y", main = "Truth")
plot(X, col = col_fit, pch = 19, cex = 0.5,
     xlab = "x", ylab = "y", main = "CAVI fit")
par(op)
```

The fitted object also provides plots of the objective and posterior mean
trajectory.

```{r plot-trace, fig.cap="CAVI trace plot from the fitted mpcurve object."}
plot(fit, plot_type = "elbo")
```

```{r plot-mu, fig.cap="Posterior mean trajectories across the K bins."}
plot(fit, plot_type = "mu")
```

## 7  Using the results

The ordering-recovery statistic above compares estimated sample positions
with the generating latent variable. Its absolute value accounts for the
arbitrary direction of an inferred ordering. In an application, use known
sample annotations to help interpret that direction and assess whether
nearby samples have similar feature profiles.

Normalized positions are available in `fit$locations$mean$pseudotime`,
trajectory means in `fit$params$mu`, and trajectory covariance matrices in
`fit$fit$posterior$cov`. If the objective is still increasing at the iteration
limit, continue with `fit <- do_mpcurve(fit, iter = 100)`.

When different feature groups may follow different sample orderings, see the
[feature-partitioning tutorial](partition.html). For data with supplied
measurement uncertainty, see the [fitness analysis](fitness.html).
