---
title: "Feature Partitioning with MPCurver"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Feature Partitioning with MPCurver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

```{r load}
library(MPCurver)
set.seed(42)
```

## The feature-partition problem

The single-ordering model assumes that all features share one latent sample
ordering. The partition model instead uses a fixed number $M$ of candidate
orderings and a feature-level assignment
$Z_j \in \{1,\ldots,M\}$. If feature $j$ belongs to ordering $m$,

$$
x_{ij} \mid Z_j=m,\; C_i^{(m)}=k,\; U_j
\sim N(u_{jk},\sigma_j^2).
$$

The feature variance $\sigma_j^2$ is shared across orderings. Each conditional
trajectory has an ordering-specific smoothness parameter,

$$
U_j \mid Z_j=m
\sim N\!\left(0,(\lambda_{jm}Q_K)^{-1}\right).
$$

The public interface uses `intrinsic_dim = M`. Version 0.3.0 supports fitting a
user-specified fixed $M$; automatic greedy selection of $M$ is temporarily
unavailable because the fixed-$M$ variational objective is not itself a
complexity-penalized model-selection criterion.

## Structural variational family

For every $M \geq 2$, MPCurver uses

$$
q(C,Z,U)
=
\prod_{m=1}^M\prod_{i=1}^n q(C_i^{(m)})
\prod_{j=1}^d q(Z_j)q(U_j\mid Z_j).
$$

Write

$$
q(C_i^{(m)}=k)=r_{imk},\qquad
q(Z_j=m)=w_{jm},
$$

and

$$
q(U_j\mid Z_j=m)=N(\mu_{jm},V_{jm}).
$$

This dependence is important: the $M$ Gaussian factors for feature $j$ are
conditional alternatives under $Z_j$, rather than $M$ independent potential
trajectories that all exist simultaneously.

### Conditional trajectory update

Let $N_{mk}=\sum_i r_{imk}$. With estimated feature variance,

$$
A_{jm}
=
\lambda_{jm}Q_K
+\operatorname{diag}\!\left(\frac{N_m}{\sigma_j^2}\right),
\qquad
b_{jm}
=
\frac{(R^{(m)})^T X_{\cdot j}}{\sigma_j^2},
$$

and

$$
V_{jm}=A_{jm}^{-1},\qquad \mu_{jm}=V_{jm}b_{jm}.
$$

The feature-assignment probability $w_{jm}$ does **not** occur in either the
precision or the right-hand side. Therefore a small or zero $w_{jm}$ does not
drive the conditional posterior back toward a singular intrinsic prior.

If known observation-level variances $v_{ij}$ are supplied through `S`, the
likelihood terms become

$$
A_{jm}
=
\lambda_{jm}Q_K
+\operatorname{diag}_k\sum_i\frac{r_{imk}}{v_{ij}},
\qquad
b_{jmk}=\sum_i\frac{r_{imk}x_{ij}}{v_{ij}}.
$$

### Ordering responsibilities

The ordering-specific sample responsibilities still average over feature
assignments:

$$
\log r_{imk}
\stackrel{+}{=}
\log\pi_{mk}
-\frac12\sum_j w_{jm}
\frac{(x_{ij}-\mu_{jmk})^2+V_{jm,kk}}{\sigma_j^2}.
$$

Thus $w_{jm}$ affects how strongly feature $j$ informs ordering $m$, while it
does not change the definition of the conditional trajectory posterior.

### Shared feature variance

For estimated noise, define

$$
E_{jm}
=
\sum_{i,k}r_{imk}
\left[(x_{ij}-\mu_{jmk})^2+V_{jm,kk}\right].
$$

The single shared variance update is

$$
\sigma_j^2
=
\frac1n\sum_{m=1}^M w_{jm}E_{jm}.
$$

Consequently, a partition fit exposes one length-$d$ vector in
`fit$params$sigma2`. When `S` is known, this field is `NULL`.

### Feature assignment and local ELBO

The feature assignment uses the complete conditional local ELBO, not only the
expected likelihood. Let

$$
G_{jm}=\mu_{jm}^TQ_K\mu_{jm}+\operatorname{tr}(Q_KV_{jm}).
$$

For an intrinsic precision with rank $r$ and generalized determinant
$|Q_K|_+$, the local block $B_{jm}$ contains the expected likelihood, the
trajectory prior, the optional lambda prior, and the Gaussian entropy:

$$
\begin{aligned}
B_{jm}
=\;&-\frac n2\log(2\pi\sigma_j^2)
-\frac{E_{jm}}{2\sigma_j^2}\\
&+\frac12\left[
r\{\log\lambda_{jm}-\log(2\pi)\}
+\log|Q_K|_+
-\lambda_{jm}G_{jm}
\right]\\
&+\log p(\lambda_{jm})
+\frac12\left[K\log(2\pi e)+\log|V_{jm}|\right].
\end{aligned}
$$

At temperature $T$,

$$
w_{jm}
\propto
\exp\left\{
\frac{\mathbb E\log\omega_m+B_{jm}}{T}
\right\}.
$$

At $T=1$, `objective_history` records the fixed-$M$ structural ELBO. Objective
values at different annealing temperatures are not directly comparable; only
the fixed-$T$ segment should be checked for coordinate-ascent monotonicity.

## Intrinsic prior with `ridge = 0`

For an intrinsic random-walk prior, $Q_K$ is rank deficient. The conditional
posterior precision is nevertheless

$$
\lambda_{jm}Q_K+D_{jm},
$$

where $D_{jm}$ is supplied by the likelihood and does not vanish with
$w_{jm}$. This is the same posterior-identification mechanism used by the
single-ordering CAVI fit. The algorithm does not use feature-weight floors,
automatic precision jitter, active/inactive orderings, or freeze/drop states.

`ridge` remains available as an explicit modelling choice. A positive value
makes the trajectory prior proper, but it is not required merely to rescue the
conditional posterior from a small feature-assignment weight.

With `ridge = 0`, the trajectory prior normalization uses the rank and
generalized determinant of $Q_K$. Accordingly, the recorded value follows the
package's intrinsic or pseudo-ELBO convention; structural VI fixes the
posterior singularity caused by the old fractional update, but it does not turn
an improper intrinsic prior into a proper marginal likelihood.

## Fitting a fixed number of orderings

```{r simulate}
sim <- simulate_dual_trajectory(
  n = 500,
  d1 = 10,
  d2 = 10,
  d_noise = 0,
  sigma = 0.05,
  trajectory_family = c("quadratic", "monotone"),
  seed = 1
)

X <- sim$X
truth <- sim$true_assign
```

```{r fit-soft}
fit_partition <- fit_mpcurve(
  X,
  intrinsic_dim = 2,
  method = "PCA",
  K = 50,
  rw_q = 2,
  ridge = 0,
  verbose = FALSE
)
```

The returned `mpcurve` object has one canonical structural state:

- `params$sigma2`: shared length-$d$ variance vector, or `NULL` with known `S`;
- `gamma`: named list of $M$ responsibility matrices;
- `conditional_posterior`: conditional means and covariances for every ordering;
- `lambda_mat`: $d\times M$ smoothness matrix;
- `partition$pi_weights`: $d\times M$ feature-assignment probabilities;
- `partition$assign`: hard labels obtained from the largest probability;
- `objective_history` and `temperature_history`: fitting diagnostics.

`fits` remains available only as a derived compatibility view for existing
read-only code. These child objects are regenerated from the canonical state;
they are not independent fitting states and are not used by `do_mpcurve()`.

```{r visualize-mpcurve, fig.cap="Inspecting the fixed-M structural objective and the ordering-specific pseudotime views."}
plot(fit_partition, plot_type = "elbo")
plot(fit_partition, dims = 4)
```

### Plotting structural trajectory means

For a partition fit, `plot_type = "mu"` reads the canonical conditional
posterior directly. With one selected feature, panel $m$ displays

$$
E[U_j \mid Z_j=m]
$$

against normalized latent position for every fixed ordering $m$. The panel
title reports $q(Z_j=m)$, which helps distinguish a well-supported conditional
trajectory from a low-probability alternative. The trajectory itself is not
multiplied by this assignment probability: it is the mean of the conditional
Gaussian branch.

```{r structural-mu-one-dimensional, fig.cap="Conditional posterior mean trajectory for feature 4 under each of the two fixed orderings."}
plot(fit_partition, plot_type = "mu", dims = 4)
```

When `dims` contains two features, each panel instead plots the two conditional
mean coordinates against one another along the latent grid. This is useful for
comparing the geometric paths implied by the same pair of features under the
different ordering assignments.

```{r structural-mu-two-dimensional, fig.cap="Two-dimensional conditional posterior mean paths for features 4 and 14 under each fixed ordering."}
plot(fit_partition, plot_type = "mu", dims = c(4, 14))
```

Both forms use `conditional_posterior$mean`, not an independently updated child
fit. Consequently, the displayed paths are exactly the
$q(U_j\mid Z_j=m)$ means used by the structural coordinate updates.

### Inspecting the fixed-temperature objective

```{r check-objective}
t1 <- fit_partition$temperature_history == 1
delta_t1 <- diff(fit_partition$objective_history[t1])
cat("Largest T=1 decrease:", min(c(0, delta_t1)), "\n")
```

### Partition accuracy

```{r accuracy}
pred <- as.character(fit_partition$partition$assign)
truth <- as.character(truth)
accuracy <- max(mean(pred == truth), mean(pred != truth))
cat("Partition accuracy:", round(accuracy, 4), "\n")
```

### Continuing a fit

```{r continue, eval=FALSE}
fit_partition <- do_mpcurve(fit_partition, iter = 10)
```

Continuation uses only the canonical structural state and appends exact
$T=1$ coordinate sweeps to `objective_history`.

## Initialization

The default `partition_init = "similarity"` clusters features and initializes
one ordering per feature block. Alternatively,
`partition_init = "ordering_methods"` accepts one method per ordering:

```{r fit-methods, eval=FALSE}
fit3 <- fit_mpcurve(
  X,
  intrinsic_dim = 3,
  partition_init = "ordering_methods",
  method = c("fiedler", "PCA", "PCA"),
  K = 16
)
```

For estimated noise, `sigma2_init` is a scalar or length-$d$ shared variance
vector. `lambda` initializes all entries of `lambda_mat`; use
`fix_lambda = TRUE` to keep them fixed. `sigma2_init` cannot be combined with
known measurement standard deviations `S`.

## Deprecated transition arguments

The former augmented algorithm exposed `freeze_unused_ordering`,
`freeze_feature`, their thresholds, and `drop_unused_ordering`. They are no
longer part of the fitting algorithm. During the 0.3.x transition they remain
accepted by `fit_mpcurve()` and `do_mpcurve()` only to issue a deprecation
warning; their values are ignored and new objects do not store active, frozen,
effective-weight, or dropped-ordering state.

## Practical guidance

- Choose a fixed, scientifically plausible `intrinsic_dim` and compare the
  resulting partitions through stability or external validation.
- Do not use the raw structural ELBO alone to select $M$: with an adaptive
  ordering prior, an extra nearly empty ordering is not automatically charged
  a model-complexity penalty.
- `ridge = 0` is the default intrinsic-prior path. Use positive `ridge` only
  when a properized trajectory prior is part of the intended model.
- Inspect only fixed-$T$ segments for monotonicity.
- Use multiple initializations when trajectories are non-monotone or weakly
  separated.

## Summary

- All fixed $M\ge2$ fits use structural VI, $q(C)q(Z)q(U\mid Z)$.
- Conditional trajectory precision is independent of $w_{jm}$.
- Estimated $\sigma_j^2$ is shared across orderings.
- New fits have one canonical state; `$fits` is a derived compatibility view.
- Freeze/drop machinery is retired, and greedy cross-$M$ selection is
  temporarily unavailable.
