---
title: "Multiple Orderings with MPCurver"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Multiple Orderings with MPCurver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

## Intrinsic dimension

A single ordering describes one axis of variation: every feature follows a
smooth trajectory along the same sample positions. Some datasets contain
several such axes. For example, genes associated with cell differentiation
and genes associated with the cell cycle can vary across the same cells in
different ways. One ordering may describe one group well while missing the
structure in the other.

MPCurver represents these patterns using $M$ latent sample orderings. Each
feature follows one ordering, and the model learns its assignment together
with the sample positions and smooth trajectory. In this model, intrinsic
dimension refers to the number of latent orderings in the returned model. With $M=1$,
all features share one ordering; with $M=2$, two feature groups can follow
different sample positions.

We first describe the model, then fit an example with $M=2$.
[Estimating intrinsic dimension](intrinsic_dimension.html) covers choosing $M$;
[Multi-ordering CAVI](cavi_partition.html) describes the algorithm separately.

## Statistical model

The standard model fits a fixed number $M$ of latent orderings to a
sample-by-feature matrix $X\in\mathbb{R}^{n\times P}$. Each ordering uses
$K$ grid positions. Sample $i$ has a position $C_i^{(m)}$ along every ordering
$m$, while feature $j$ follows one ordering, indicated by $Z_j$:

$$
\Pr(C_i^{(m)}=k)=\pi_{mk},\qquad
\Pr(Z_j=m)=\omega_m.
$$

The position probabilities sum to one over $k$, and the feature-assignment
prior weights sum to one over $m$. A feature's assignment is shared across
all samples.

### Observations and smooth trajectories

Write $U_j=(u_{j1},\ldots,u_{jK})^T$ for feature $j$'s trajectory. Conditional
on its ordering and the sample's position,

$$
x_{ij}\mid Z_j=m,\ C_i^{(m)}=k,\ U_j
\sim N(u_{jk},\sigma_j^2).
$$

There is one noise variance $\sigma_j^2$ per feature, shared across orderings
and grid positions. The trajectory has a random-walk smoothness prior:

$$
p(U_j\mid Z_j=m,\lambda_{jm})
\propto
\exp\!\left(-\frac{\lambda_{jm}}{2}U_j^TQ_KU_j\right).
$$

The default second-order random walk (RW2) penalizes second differences along
the grid. A larger precision $\lambda_{jm}$ favors a smoother trajectory.
The standard fit estimates the noise variances, smoothness precisions,
position probabilities $\pi$, and feature-assignment prior weights $\omega$
by empirical Bayes.

With the default `ridge = 0`, $Q_K$ is an intrinsic precision matrix of rank
$K-2$. The likelihood identifies the directions left unpenalized by the prior.

The fitting algorithm learns sample positions and soft feature-assignment
probabilities jointly, retaining a trajectory distribution conditional on each
possible assignment. The [Multi-ordering CAVI](cavi_partition.html) page gives
the variational family, objective contributions, and coordinate updates.

## Choose a fitting approach

When the dimension is unknown, start with adaptive fitting using
`fit_mpcurve(X, intrinsic_dim = "auto")`. It is computationally simpler:
similarity initialization chooses a starting dimension, then one adaptive
model is fitted and orderings with numerical-zero estimated prior weights are
removed. It does not require a sequence of full fits at different dimensions.

- Use `intrinsic_dim = M` when the analysis specifies the number of orderings.
- Use `intrinsic_dim = "auto"` with the default adaptive partition prior when
  you want the model to estimate the number of supported orderings.
- Use `select_mpcurve_dimension()` when you specifically want to compare
  dimensions under a fixed uniform partition prior. It fits multiple candidate
  models and compares their final objectives.

The adaptive and fixed-uniform procedures use different priors and selection
rules, so they may return different dimensions. See
[Estimating intrinsic dimension](intrinsic_dimension.html) for examples.

## Fit a model with M = 2

```{r load}
library(MPCurver)
set.seed(42)
```

We simulate 500 samples with two groups of ten features. The groups vary along
different latent orderings, with quadratic and monotone trajectory families.

```{r simulate}
sim <- simulate_dual_trajectory(
  n = 500,
  d1 = 10,
  d2 = 10,
  d_noise = 0,
  noise_sd = 0.05,
  trajectory_family = c("quadratic", "monotone"),
  seed = 1
)

X <- sim$X
truth <- sim$true_assign
```

Fit a model with $M=2$ and 50 grid positions using the default RW2 prior
and PCA initialization.

```{r fit-soft}
fit_partition <- fit_mpcurve(
  X,
  intrinsic_dim = 2,
  num_bins = 50
)
```

### Interpret feature assignments

`fitted_assignments(fit_partition)` is a feature-by-ordering matrix:
entry $(j,m)$ is the fitted probability $q(Z_j=m)$ that feature $j$ follows
ordering $m$. Each row sums to one. A row concentrated on one ordering gives
a clear assignment; a diffuse row indicates that several orderings describe
the feature similarly well.

```{r assignment-probabilities}
head(fitted_assignments(fit_partition))
table(fit_partition$partition$assign)
```

The labels in `partition$assign` select the most probable ordering for each
feature. For this simulation we can compare those labels with the generating
groups, allowing for an exchange of the two ordering labels.

```{r accuracy}
pred <- as.character(fit_partition$partition$assign)
truth <- as.character(truth)
accuracy <- max(mean(pred == truth), mean(pred != truth))
cat("Partition accuracy:", round(accuracy, 4), "\n")
```

### Inspect sample orderings and trajectories

`plot(fit_partition, dims = 4)` displays the observed values of feature 4
against the inferred sample positions for each ordering, together with its
fitted trajectory. The objective plot shows progress over fitting iterations.

Figure 1 shows the variational objective over fitting iterations.
Figure 2 shows observed feature values along each inferred ordering.

```{r visualize-mpcurve, fig.cap="Figure 1. Variational objective over fitting iterations."}
plot(fit_partition, plot_type = "elbo")
```

```{r visualize-sample-orderings, fig.cap="Figure 2. Observed feature 4 values along each inferred ordering; colors encode pseudotime, and orange lines show fitted mean trajectories."}
plot(fit_partition, dims = 4, add_legend = FALSE,
     pal = hcl.colors(100, "Viridis"))
```

To focus on trajectory shape, use `plot_type = "mu"`. Panel $m$ displays the
conditional posterior mean $E[U_j\mid Z_j=m]$ against normalized latent
position. The assignment probability in the panel title indicates how
strongly the data support that ordering for the selected feature.

Figure 3 shows conditional mean trajectory of feature 4 under each ordering; panel titles report assignment probabilities.

```{r structural-mu-one-dimensional, fig.cap="Figure 3. Conditional mean trajectory of feature 4 under each ordering; panel titles report assignment probabilities."}
plot(fit_partition, plot_type = "mu", dims = 4)
```

Each curve answers: what trajectory is estimated if this feature follows this
ordering? A low-probability panel describes an alternative assignment and
should be interpreted with that probability in view.

With two selected features, each panel traces their conditional means against
one another across the latent grid. This shows the joint path implied by
placing both features on that ordering. Check both assignment probabilities
when interpreting the path.

Figure 4 shows conditional mean paths for features 4 and 14 under each ordering.

```{r structural-mu-two-dimensional, fig.cap="Figure 4. Conditional mean paths for features 4 and 14 under each ordering."}
plot(fit_partition, plot_type = "mu", dims = c(4, 14))
```

### Extract results

Table 1 lists the extraction functions used for further analysis. Their
ordering dimension is retained for both single- and multiple-ordering fits.

Table 1. Result extraction for sample positions, conditional trajectories, and feature assignments. Here $n$ is the number of samples, $P$ features, $K$ bins, and $M$ fitted orderings.

| Quantity | Extraction | Shape |
|:--|:--|:--|
| Posterior-mean sample positions | `fitted_positions(fit_partition)` | $n \times M$ |
| Position probabilities | `fitted_positions(fit_partition, "probability")` | $n \times K \times M$ |
| Feature-assignment probabilities | `fitted_assignments(fit_partition)` | $P \times M$ |
| Conditional trajectory means | `fitted_trajectories(fit_partition)` | $P \times K \times M$ |
| Conditional trajectory SDs | `fitted_trajectories(fit_partition, "sd")` | $P \times K \times M$ |
| Conditional trajectory covariances | `fitted_trajectories(fit_partition, "covariance")` | $K \times K \times P \times M$ |

```{r extract-results}
positions <- fitted_positions(fit_partition)
assignment_probabilities <- fitted_assignments(fit_partition)
trajectories <- fitted_trajectories(fit_partition)
```

Rows retain the input sample or feature names. Use feature names in plots,
for example `plot(fit_partition, dims = colnames(X)[1])` when `X` has names.
Model metadata such as `fit_partition$intrinsic_dim` and
`fit_partition$params$sigma2` remain available directly.

Sample positions are normalized to $[0,1]$. Ordering labels and directions
are arbitrary; use sample annotations to interpret the axes and align
orderings when comparing fits. See `?mpcurve` for the complete object
description. Effective-dimension reporting is covered in
[Estimating intrinsic dimension](intrinsic_dimension.html#report-the-supported-dimension).

### Assess convergence and continue fitting

By default, fitting runs directly at temperature $T=1$, with no annealing.
`max_iter` limits the coordinate sweeps, and `objective_history` records
improvements in the same objective throughout fitting.

```{r check-objective}
objective <- fit_partition$objective_history
cat("Recorded objective values:", length(objective), "\n")
cat("Final objective increments:\n")
tail(diff(objective))
```

If the increments remain appreciable at the iteration limit, run additional
iterations:

```{r continue, eval=FALSE}
fit_partition <- do_mpcurve(fit_partition, max_iter = 10)
```

This refines the orderings, trajectories, assignments, and estimated
parameters at $T=1$, extending the objective history.

### Fitting settings {#choose-fitting-settings}

#### Initialization

For a specified $M$, the default initialization uses feature similarity and
single-linkage clustering to form $M$ feature groups, then initializes one
sample ordering per group. Each group uses the selected `initial_method`;
the default `"PCA"` uses that group's first principal component.

Similarity options are `"spearman"` (the default for specified $M$),
`"pearson"`, `"spline_r2"`, and `"smooth_fit"`. For example, pass
`init_control = mpcurve_init_control(similarity_metric = "spline_r2")`
to change the metric. See
[`mpcurve_init_control()`](../reference/mpcurve_init_control.html#feature-similarity)
for metric definitions and alternative initialization settings.

#### Noise and smoothness

By default, the model estimates the noise variance $\sigma_j^2$ for each
feature $j$, shared across orderings. `control$sigma2_init` sets the initial
values of $\sigma_j^2$ using a scalar or a feature-specific vector.
Alternatively, supply known measurement standard
deviations through `S`, as a feature-specific vector or a matrix matching
`X`. In that case `params$sigma2` is `NULL`, and the supplied
measurement errors determine the likelihood variances.

See [the known-error updates](cavi_partition.html#known-measurement-errors)
for the corresponding algorithm.

For advanced fitting settings, see
[`mpcurve_control()`](../reference/mpcurve_control.html). To estimate the
intrinsic dimension, continue to [Estimating intrinsic dimension](intrinsic_dimension.html).
