---
title: "Feature Partitioning with MPCurver"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Feature Partitioning with MPCurver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

```{r load}
library(MPCurver)
set.seed(42)
```

## When to use multiple orderings

Different groups of features can reflect different patterns among the same
samples. For example, one group of genes may track cell differentiation while
another tracks a separate cell-state axis. A feature-partition model estimates
several sample orderings and the probability that each feature follows each
one.

Set `intrinsic_dim = M` to specify the number of orderings. The fit returns
sample positions along each ordering, smooth feature trajectories conditional
on each assignment, and assignment probabilities that quantify support for
the feature groups.

This tutorial starts with a two-ordering example, shows how to interpret its
results, and then explains the model and structural variational updates.

## Fit a two-ordering model

We simulate 500 samples with two groups of ten features. The groups vary along
different latent orderings, with quadratic and monotone trajectory families.

```{r simulate}
sim <- simulate_dual_trajectory(
  n = 500,
  d1 = 10,
  d2 = 10,
  d_noise = 0,
  sigma = 0.05,
  trajectory_family = c("quadratic", "monotone"),
  seed = 1
)

X <- sim$X
truth <- sim$true_assign
```

Fit two orderings using 50 grid positions and a second-order random-walk
prior. The default initialization groups similar features and uses PCA to
initialize a sample ordering within each group.

```{r fit-soft}
fit_partition <- fit_mpcurve(
  X,
  intrinsic_dim = 2,
  method = "PCA",
  K = 50,
  rw_q = 2,
  ridge = 0,
  verbose = FALSE
)
```

### Interpret feature assignments

`fit_partition$partition$pi_weights` is a feature-by-ordering matrix:
entry $(j,m)$ is the fitted probability $q(Z_j=m)$ that feature $j$ follows
ordering $m$. Each row sums to one. A row concentrated on one ordering gives
a clear assignment; a diffuse row indicates that several orderings describe
the feature similarly well.

```{r assignment-probabilities}
head(fit_partition$partition$pi_weights)
table(fit_partition$partition$assign)
```

The labels in `partition$assign` select the most probable ordering for each
feature. For this simulation we can compare those labels with the generating
groups, allowing for an exchange of the two ordering labels.

```{r accuracy}
pred <- as.character(fit_partition$partition$assign)
truth <- as.character(truth)
accuracy <- max(mean(pred == truth), mean(pred != truth))
cat("Partition accuracy:", round(accuracy, 4), "\n")
```

### Inspect sample orderings and trajectories

`plot(fit_partition, dims = 4)` displays the observed values of feature 4
against the inferred sample positions for each ordering, together with its
fitted trajectory. The objective plot shows progress over fitting iterations.

```{r visualize-mpcurve, fig.cap="Variational objective over fitting iterations and observed feature values along each inferred ordering."}
plot(fit_partition, plot_type = "elbo")
plot(fit_partition, dims = 4)
```

To focus on trajectory shape, use `plot_type = "mu"`. Panel $m$ displays the
conditional posterior mean $E[U_j\mid Z_j=m]$ against normalized latent
position. The assignment probability in the panel title indicates how
strongly the data support that ordering for the selected feature.

```{r structural-mu-one-dimensional, fig.cap="Conditional mean trajectory of feature 4 under each ordering; panel titles report assignment probabilities."}
plot(fit_partition, plot_type = "mu", dims = 4)
```

Each curve answers: what trajectory is estimated if this feature follows this
ordering? A low-probability panel describes an alternative assignment and
should be interpreted with that probability in view.

With two selected features, each panel traces their conditional means against
one another across the latent grid. This shows the joint path implied by
placing both features on that ordering. Check both assignment probabilities
when interpreting the path.

```{r structural-mu-two-dimensional, fig.cap="Conditional mean paths for features 4 and 14 under each ordering."}
plot(fit_partition, plot_type = "mu", dims = c(4, 14))
```

### Extract results

The following fields support further analysis:

| Quantity | Field |
|:--|:--|
| Sample positions for ordering $m$ | `locations[[m]]$mean$pseudotime` |
| Feature-assignment probabilities | `partition$pi_weights` |
| Most probable feature assignments | `partition$assign` |
| Conditional trajectory means | `conditional_posterior$mean[[m]]` |
| Trajectory covariance for feature $j$, ordering $m$ | `conditional_posterior$cov[[m]][[j]]` |
| Feature noise variances, shared across orderings | `params$sigma2` |
| Feature-by-ordering smoothness parameters | `lambda_mat` |

Sample positions are normalized to $[0,1]$. Ordering labels and directions
are arbitrary; use sample annotations to interpret the axes and align
orderings when comparing fits. See `?mpcurve` for the complete object
description.

### Assess convergence and continue fitting

Fitting begins with an annealing schedule that encourages diffuse feature
assignments, followed by coordinate updates at temperature $T=1$. Inspect
`objective_history` together with `temperature_history`: the $T=1$ segment
tracks improvements in a common objective. Values from different
temperatures correspond to different objectives.

```{r check-objective}
t1_objective <- fit_partition$objective_history[
  fit_partition$temperature_history == 1
]
cat("Recorded T=1 objective values:", length(t1_objective), "\n")
cat("Final objective increments:\n")
tail(diff(t1_objective))
```

If the increments remain appreciable at the iteration limit, run additional
iterations:

```{r continue, eval=FALSE}
fit_partition <- do_mpcurve(fit_partition, iter = 10)
```

This refines the orderings, trajectories, assignments, and estimated
parameters at $T=1$, extending the objective history.

## Choose fitting settings

### Number of orderings

`intrinsic_dim` determines the number of orderings and stays fixed during
fitting. Choose values that address the scientific question, then examine
the stability of feature groups and their interpretation across fits.
The structural objective assesses convergence for a given $M$; it does not
include a complexity penalty for comparing different values of $M$.
Use a separate predictive or model-selection assessment for that comparison.

### Initialization

The default `partition_init = "similarity"` clusters features by similarity
and initializes one ordering per feature group. Alternatively,
`partition_init = "ordering_methods"` uses the specified ordering methods:

```{r fit-methods, eval=FALSE}
fit3 <- fit_mpcurve(
  X,
  intrinsic_dim = 3,
  partition_init = "ordering_methods",
  method = c("fiedler", "PCA", "PCA"),
  K = 16
)
```

Compare several initializations when trajectories are non-monotone or
groups are weakly separated. Within a common model and fitting configuration,
compare final $T=1$ objective values as well as the resulting orderings and
feature assignments.

### Noise and smoothness

By default, the model estimates one noise variance per feature, shared
across orderings. `sigma2_init` supplies a scalar or a vector of initial
feature variances. Alternatively, supply known measurement standard
deviations through `S`, as a feature-specific vector or a matrix matching
`X`. In that case `params$sigma2` is `NULL`, and the supplied
measurement errors determine the likelihood variances.

`lambda` initializes the smoothness parameters. Larger values give
stronger smoothing; `fix_lambda = TRUE` holds them fixed during fitting.
An optional `lambda_sd_prior_rate` encourages smooth trajectories while
allowing their smoothness parameters to be estimated.

## Statistical model

Let $C_i^{(m)}$ be the latent grid position of sample $i$ along ordering $m$,
and let $Z_j\in\{1,\ldots,M\}$ identify the ordering followed by feature $j$.
The position and feature-assignment probabilities are
$\Pr(C_i^{(m)}=k)=\pi_{mk}$ and $\Pr(Z_j=m)=\omega_m$. Conditional on these
assignments,

$$
x_{ij}\mid Z_j=m,\; C_i^{(m)}=k,\; U_j
\sim N(u_{jk},\sigma_j^2).
$$

Each feature has a trajectory $U_j$ whose prior smoothness depends on its
ordering:

$$
p(U_j\mid Z_j=m,\lambda_{jm})
\propto
\exp\!\left(-\frac{\lambda_{jm}}{2}U_j^TQ_KU_j\right).
$$

$Q_K$ is the random-walk precision matrix. Its quadratic form penalizes
variation along the ordered grid. With known measurement standard deviations,
replace $\sigma_j^2$ by $v_{ij}=S_{ij}^2$ (or $S_j^2$ for a feature-specific
vector).

### Structural variational inference

MPCurver approximates the posterior by

$$
q(C,Z,U)
=
\prod_{m=1}^M\prod_{i=1}^n q(C_i^{(m)})
\prod_{j=1}^d q(Z_j)q(U_j\mid Z_j).
$$

Write $q(C_i^{(m)}=k)=r_{imk}$, $q(Z_j=m)=w_{jm}$, and

$$
q(U_j\mid Z_j=m)=N(\mu_{jm},V_{jm}).
$$

The conditional Gaussian distributions describe the possible trajectory
shapes under each feature assignment. Preserving this dependence allows
assignment uncertainty and trajectory uncertainty to be represented together.
Coordinate updates alternate between trajectories, sample positions, feature
assignments, and model parameters.

### Conditional trajectory update

Let $R^{(m)}=(r_{imk})$ and $N_{mk}=\sum_i r_{imk}$. With estimated feature
variance,

$$
A_{jm}
=
\lambda_{jm}Q_K
+\operatorname{diag}\!\left(\frac{N_m}{\sigma_j^2}\right),
\qquad
b_{jm}
=
\frac{(R^{(m)})^T X_{\cdot j}}{\sigma_j^2},
$$

and

$$
V_{jm}=A_{jm}^{-1},\qquad \mu_{jm}=V_{jm}b_{jm}.
$$

These updates combine the smoothness prior with the full likelihood
conditional on $Z_j=m$. Thus each possible assignment has its own trajectory
estimate and covariance, including assignments with small $w_{jm}$.

With known variances $v_{ij}$, the likelihood terms become

$$
A_{jm}
=
\lambda_{jm}Q_K
+\operatorname{diag}_k\sum_i\frac{r_{imk}}{v_{ij}},
\qquad
b_{jmk}=\sum_i\frac{r_{imk}x_{ij}}{v_{ij}}.
$$

### Ordering responsibilities

The sample-position update weights each feature's contribution by its
assignment probability:

$$
\log r_{imk}
\stackrel{+}{=}
\log\pi_{mk}
-\frac12\sum_j w_{jm}
\frac{(x_{ij}-\mu_{jmk})^2+V_{jm,kk}}{\sigma_j^2},
$$

where $\stackrel{+}{=}$ denotes equality up to a normalizing constant in $k$.
A feature with large $w_{jm}$ strongly informs ordering $m$. The term
$V_{jm,kk}$ propagates trajectory uncertainty into the position update.
For known measurement variances, replace $\sigma_j^2$ by $v_{ij}$.

### Shared feature variance

For estimated noise, define

$$
E_{jm}
=
\sum_{i,k}r_{imk}
\left[(x_{ij}-\mu_{jmk})^2+V_{jm,kk}\right].
$$

The variance update averages the expected squared residuals across ordering
assignments:

$$
\sigma_j^2
=
\frac1n\sum_{m=1}^M w_{jm}E_{jm}.
$$

This gives one noise scale for each feature. When measurement variances are
supplied through `S`, those variances are used directly.

### Feature assignments

Assignment probabilities balance expected data fit, trajectory smoothness,
and posterior uncertainty. Define

$$
G_{jm}=\mu_{jm}^TQ_K\mu_{jm}+\operatorname{tr}(Q_KV_{jm}).
$$

With estimated noise, precision rank $r$, and generalized determinant
$|Q_K|_+$, the local objective for feature $j$ under ordering $m$ is

$$
\begin{aligned}
B_{jm}
=\;&-\frac n2\log(2\pi\sigma_j^2)
-\frac{E_{jm}}{2\sigma_j^2}\\
&+\frac12\left[
r\{\log\lambda_{jm}-\log(2\pi)\}
+\log|Q_K|_+
-\lambda_{jm}G_{jm}
\right]\\
&+\log p(\lambda_{jm})
+\frac12\left[K\log(2\pi e)+\log|V_{jm}|\right].
\end{aligned}
$$

The terms are the expected likelihood, expected trajectory log prior,
optional smoothness-parameter log prior, and Gaussian entropy. The default
smoothness-parameter penalty is zero. At annealing temperature $T$,

$$
w_{jm}
\propto
\exp\left\{
\frac{\log\omega_m+B_{jm}}{T}
\right\}
$$

for a fixed or empirically estimated ordering prior $\omega$.
At $T=1$, these are the feature-assignment updates for the structural
variational objective.

### Intrinsic random-walk prior

The default `ridge = 0` gives an intrinsic random-walk prior. $Q_K$ is rank
deficient, and the likelihood provides information about the unpenalized
directions. The conditional posterior is proper when
$\lambda_{jm}Q_K+D_{jm}$ is positive definite, equivalently when

$$
\ker(Q_K)\cap\ker(D_{jm})=\{0\},
$$

where $D_{jm}$ is the diagonal likelihood precision in the trajectory update.
This is the same identification condition as in the single-ordering model.
A positive `ridge` adds a diagonal term to $Q_K$ and gives a proper Gaussian
trajectory prior.

For an intrinsic prior, the objective uses the rank and generalized
determinant of $Q_K$, with a fixed reference measure on the null space. It
supports fitting and convergence assessment for the specified model; an
ordinary marginal-likelihood interpretation requires a proper prior.
