---
title: "Mathematical Details of the Single-Ordering MPCurver CAVI Algorithm"
output:
  rmarkdown::html_vignette:
    toc: true
    mathjax: default
vignette: >
  %\VignetteIndexEntry{Mathematical Details of the Single-Ordering MPCurver CAVI Algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load}
library(MPCurver)
```

This note gives the supplementary-methods derivation for the current
single-ordering `cavi()` implementation.  It covers both estimated feature
variance and known measurement standard deviations, proper and intrinsic
random-walk precisions, fixed and adaptive position priors, and the optional
prior on trajectory scale.

The fixed-$M$ multi-ordering extension is derived separately in
`partition_coherent_elbo_derivation.Rmd`.  Its variational family is
$q(C)\prod_jq(Z_j)q(U_j\mid Z_j)$.  The present note concerns its mathematical
$M=1$ reduction, for which no feature-assignment variable is needed.

## 1  Notation and model

Let

- $i=1,\ldots,n$ index observations;
- $j=1,\ldots,d$ index features; and
- $k=1,\ldots,K$ index ordered mixture positions.

The observed matrix is $X=(x_{ij})\in\mathbb R^{n\times d}$.  Each observation
has a latent position

\[
C_i\in\{1,\ldots,K\},
\qquad
C_i\mid\pi\sim\operatorname{Categorical}(\pi),
\qquad
\sum_{k=1}^K\pi_k=1.
\]

Feature $j$ has a latent trajectory

\[
U_j=(U_{j1},\ldots,U_{jK})^\top\in\mathbb R^K.
\]

### 1.1  Estimated feature variance

When `S = NULL`, the observation model is

\[
x_{ij}\mid C_i=k,U_j,\sigma_j^2
\sim N(U_{jk},\sigma_j^2).
\]

There is one fitted variance $\sigma_j^2$ for feature $j$, shared across all
observations and positions.

### 1.2  Known measurement standard deviations

When `S` is supplied, it contains standard deviations.  A length-$d$ vector
specifies $s_{ij}=s_j$ for every observation; an $n\times d$ matrix specifies
$s_{ij}$ separately.  Then

\[
x_{ij}\mid C_i=k,U_j,S
\sim N(U_{jk},s_{ij}^2).
\]

Write $\tau_{ij}=s_{ij}^{-2}$.  Supplied values are finite and nonnegative;
the implementation applies a small standard-deviation floor before inversion.
No $\sigma_j^2$ parameter is estimated in this mode.  The full known-noise
derivation and its fixed-$M$ extension are also given in
`known_noise_cavi_derivation.Rmd`.

### 1.3  Random-walk trajectory prior

Let $Q_K$ be the unit-scale RW$(q)$ precision.  Define

\[
r_Q=\operatorname{rank}(Q_K),
\qquad
\log|Q_K|_+=\sum_{\nu_\ell>0}\log\nu_\ell.
\]

The trajectory prior contribution is represented as

\[
\log p(U_j\mid\lambda_j)
=\frac12\left\{
r_Q(\log\lambda_j-\log(2\pi))
+\log|Q_K|_+
\right\}
-\frac{\lambda_j}{2}U_j^\top Q_KU_j.
\]

With the default intrinsic prior (`ridge = 0`), $Q_K$ has nullity $q$ and
$r_Q=K-q$.  The expression above is an intrinsic log density under the
package's fixed pseudo-determinant and null-space reference convention.  It is
not a normalized density on all of $\mathbb R^K$.

If a positive ridge makes $Q_K$ full rank, then $r_Q=K$,
$|Q_K|_+$ is the ordinary determinant, and the expression is the proper
Gaussian log density.

### 1.4  Optional prior on trajectory scale

The smoothness $\lambda_j$ is a point-estimated parameter.  By default there
is no additional penalty.  If `lambda_sd_prior_rate = rho > 0`, the model
places an exponential prior on

\[
T_j=\lambda_j^{-1/2},
\qquad
T_j\sim\operatorname{Exp}(\rho).
\]

The induced log-prior contribution is

\[
g_\rho(\lambda_j)
=\log(\rho/2)-\frac32\log\lambda_j
-\rho\lambda_j^{-1/2}.
\]

When the option is `NULL`, $g_\rho$ is zero.  For backward compatibility, an
explicit numeric zero also means that the penalty is absent; it is not treated
as a literal exponential distribution with rate zero.

## 2  Variational family and posterior summaries

The single-ordering mean-field family is

\[
q(C,U)
=\prod_{i=1}^nq_i(C_i)\prod_{j=1}^dq_j(U_j).
\]

Write

\[
r_{ik}=q_i(C_i=k),
\qquad
\sum_{k=1}^Kr_{ik}=1,
\qquad
N_k=\sum_{i=1}^nr_{ik}.
\]

Each trajectory factor is Gaussian:

\[
q_j(U_j)=N(\mu_j,V_j).
\]

The second moment appearing in the trajectory prior is

\[
A_j
:=\mathbb E_q[U_j^\top Q_KU_j]
=\mu_j^\top Q_K\mu_j+\operatorname{tr}(Q_KV_j).
\]

The Gaussian entropy is

\[
\mathcal H_j
=H\{q_j(U_j)\}
=\frac12\{K\log(2\pi e)+\log|V_j|\}.
\]

The parameters $\pi$, $\lambda$, and, when present, $\sigma^2$ are optimized
as point-parameter blocks rather than assigned variational factors.

## 3  Exact objective represented by `elbo_trace`

### 3.1  Position block

The expected position-prior contribution plus categorical entropy is

\[
\mathcal C
=\sum_{i=1}^n\sum_{k=1}^Kr_{ik}\log\pi_k
-\sum_{i=1}^n\sum_{k=1}^Kr_{ik}\log r_{ik}.
\]

### 3.2  Estimated-noise likelihood

Since

\[
\mathbb E_q\{(x_{ij}-U_{jk})^2\}
=(x_{ij}-\mu_{jk})^2+V_{j,kk},
\]

the expected log likelihood is

\[
\mathcal D^{\mathrm{est}}
=-\frac12\sum_{i=1}^n\sum_{k=1}^K\sum_{j=1}^d
r_{ik}
\left[
\log(2\pi\sigma_j^2)
+\frac{(x_{ij}-\mu_{jk})^2+V_{j,kk}}{\sigma_j^2}
\right].
\]

### 3.3  Known-noise likelihood

With supplied measurement standard deviations,

\[
\mathcal D^S
=-\frac12\sum_{i=1}^n\sum_{k=1}^K\sum_{j=1}^d
r_{ik}
\left[
\log(2\pi s_{ij}^2)
+\frac{(x_{ij}-\mu_{jk})^2+V_{j,kk}}{s_{ij}^2}
\right].
\]

The observation-specific normalizing constants are retained in the recorded
objective even though they cancel from the position softmax.

### 3.4  Trajectory-prior block

For feature $j$,

\[
\mathcal P_j
=\frac12\left\{
r_Q(\log\lambda_j-\log(2\pi))
+\log|Q_K|_+
\right\}
-\frac{\lambda_j}{2}A_j.
\]

The complete objective is

\[
\boxed{
\mathcal L
=\mathcal C+\mathcal D
+\sum_{j=1}^d
\{\mathcal P_j+g_\rho(\lambda_j)+\mathcal H_j\},
}
\]

where $\mathcal D=\mathcal D^{\mathrm{est}}$ or $\mathcal D^S$ according to
the noise specification.  This is exactly the quantity represented in
`elbo_trace`, including the Gaussian normalizing constants and the optional
lambda prior.

### 3.5  ELBO versus intrinsic pseudo-ELBO

If $Q_K$ is proper, then for fixed point parameters and without changing the
optional parameter-prior interpretation,

\[
\log p(X\mid\theta)
=\operatorname{ELBO}(q;\theta)
+\operatorname{KL}\{q(C,U)\|p(C,U\mid X,\theta)\}.
\]

Thus the unpenalized $\mathcal L$ is an ordinary lower bound on the marginal
log likelihood.  Adding $g_\rho(\lambda)$ yields the corresponding penalized
or joint parameter-prior objective.

For an intrinsic $Q_K$, the prior is improper on its null space, so a
normalized $p(X\mid\theta)$ and the KL identity above are not available
without an explicit constraint or properization.  In that case $\mathcal L$
is a pseudo-ELBO under one fixed intrinsic reference convention.  Coordinate
comparisons within that convention remain meaningful, but the numeric value
must not be described as normalized model evidence.

## 4  Conditional Gaussian update

### 4.1  Estimated feature variance

Define

\[
D_j
=\operatorname{diag}\left(
N_1/\sigma_j^2,\ldots,N_K/\sigma_j^2
\right),
\]

\[
b_{jk}
=\frac1{\sigma_j^2}\sum_{i=1}^nr_{ik}x_{ij}.
\]

Completing the square gives

\[
P_j=\lambda_jQ_K+D_j,
\qquad
P_j\mu_j=b_j,
\qquad
V_j=P_j^{-1}.
\]

### 4.2  Known measurement standard deviations

Set $\tau_{ij}=s_{ij}^{-2}$.  Then

\[
D_j
=\operatorname{diag}\left(
\sum_ir_{i1}\tau_{ij},\ldots,
\sum_ir_{iK}\tau_{ij}
\right),
\]

\[
b_{jk}=\sum_{i=1}^nr_{ik}\tau_{ij}x_{ij},
\]

and the same Gaussian equations apply:

\[
P_j=\lambda_jQ_K+D_j,
\qquad
P_j\mu_j=b_j,
\qquad
V_j=P_j^{-1}.
\]

If `S` is a length-$d$ vector, these expressions reduce to
$D_{j,kk}=N_k/s_j^2$ and
$b_{jk}=s_j^{-2}\sum_ir_{ik}x_{ij}$.

### 4.3  Numerical empty-component protection

The theoretical estimated-noise coordinate uses the exact $N_k$.  In the
single-ordering implementation, values below $10^{-8}$ are replaced by
$10^{-8}$ only inside the estimated-noise Gaussian solve.  This protects the
Cholesky factorization if floating-point responsibilities create an exactly
empty position.  It is a likelihood-diagonal stabilization, not a ridge added
to $Q_K$, and it is normally inactive because finite categorical logits give
positive responsibilities.

The known-noise solve uses the precision-weighted masses directly.  If a
conditional precision is not positive definite, the implementation reports a
detailed precision failure rather than silently adding Cholesky jitter.

## 5  Position-responsibility update

### 5.1  Estimated feature variance

For observation $i$,

\[
\log r_{ik}
\stackrel{+c}{=}
\log\pi_k
-\frac12\sum_{j=1}^d
\left[
\log(2\pi\sigma_j^2)
+\frac{(x_{ij}-\mu_{jk})^2+V_{j,kk}}{\sigma_j^2}
\right].
\]

### 5.2  Known measurement standard deviations

With known noise,

\[
\log r_{ik}
\stackrel{+c}{=}
\log\pi_k
-\frac12\sum_{j=1}^d
\left[
\log(2\pi s_{ij}^2)
+\frac{(x_{ij}-\mu_{jk})^2+V_{j,kk}}{s_{ij}^2}
\right].
\]

In both cases the row is normalized by a log-sum-exp.  The posterior-variance
term $V_{j,kk}$ is the uncertainty correction that distinguishes CAVI from a
plug-in EM position update.

## 6  Point-parameter updates

### 6.1  Position prior

If `position_prior = "adaptive"`, the coordinate maximizer is

\[
\pi_k=\frac{N_k}{n}.
\]

The implementation applies a machine-precision floor and renormalizes the
result.  If `position_prior = "fixed"`, the supplied or initialized $\pi$ is
not updated.

### 6.2  Estimated feature variance

For `S = NULL`, differentiating the $\sigma_j^2$ block gives

\[
\boxed{
\sigma_j^2
=\frac1n\sum_{i=1}^n\sum_{k=1}^Kr_{ik}
\{(x_{ij}-\mu_{jk})^2+V_{j,kk}\}.
}
\]

This is projected onto `[sigma_min, sigma_max]`.  With known `S`, this block
does not exist and `params$sigma2` is `NULL`.

### 6.3  Smoothness without the optional scale prior

Let

\[
A_j=\mu_j^\top Q_K\mu_j+\operatorname{tr}(Q_KV_j).
\]

When $g_\rho=0$, the unconstrained maximizer is

\[
\boxed{
\lambda_j=\frac{r_Q}{A_j}.
}
\]

The implementation protects a numerically zero denominator and then projects
onto `[lambda_min, lambda_max]`.

### 6.4  Smoothness with the induced exponential prior

When $\rho>0$, the lambda-dependent objective is

\[
h_j(\lambda_j)
=\frac{r_Q-3}{2}\log\lambda_j
-\frac{A_j}{2}\lambda_j
-\rho\lambda_j^{-1/2}.
\]

The implementation maximizes $h_j$ in $\eta_j=\log\lambda_j$ within the
lambda bounds.  Its derivative is

\[
\frac{d h_j(e^{\eta_j})}{d\eta_j}
=\frac{r_Q-3}{2}
-\frac{A_j}{2}e^{\eta_j}
+\frac{\rho}{2}e^{-\eta_j/2},
\]

which is strictly decreasing.  The unique interior root is found when it
falls within the allowed interval; otherwise the appropriate boundary is the
maximizer.

If `fix_lambda = TRUE`, neither lambda update is performed.  A lambda value
explicitly supplied to continuation is also held fixed thereafter.

## 7  Sweep order and monotonicity

After the initialization objective is recorded, one `cavi()` sweep updates:

1. $q(C)$ from the current trajectory posterior;
2. adaptive $\pi$, if requested;
3. $\sigma^2$, only for estimated noise;
4. $\lambda$, unless fixed; and
5. $q(U)$ from the updated parameters.

The code then recomputes the complete objective $\mathcal L$.  Subject to the
explicit bounds, each mathematical block update is a coordinate maximizer of
the same proper ELBO, penalized ELBO, or intrinsic pseudo-ELBO described in
Section 3.  Therefore `elbo_trace` is the primary monotonicity diagnostic and
should be nondecreasing up to numerical tolerance.

Two numerical qualifications are important:

1. the estimated-noise Gaussian solve uses the empty-component mass floor in
   Section 4.3 if needed; and
2. finite-precision linear algebra and lower/upper bounds can produce tiny
   numerical differences.

The implementation checks every new objective value and warns if it decreases
beyond its tolerance.  By default, `do_cavi()` and single-ordering
`do_mpcurve()` append sweeps using the stored noise specification and objective
controls.  If a caller explicitly changes the lambda-prior rate during
continuation, the objective definition changes at that continuation boundary.

### 7.1  The plug-in log-likelihood is not the objective

`loglik_trace` evaluates the observed mixture likelihood after replacing each
trajectory by its posterior mean:

\[
\ell_{\mathrm{plug\mbox{-}in}}
=\sum_{i=1}^n\log
\left[
\sum_{k=1}^K\pi_k
\prod_{j=1}^d
N(x_{ij};\mu_{jk},v_{ij})
\right],
\]

where $v_{ij}=\sigma_j^2$ for estimated noise and $v_{ij}=s_{ij}^2$ for known
noise.  It omits $V_j$ and does not integrate over $U$.  It is neither the
marginal log likelihood nor the monotonicity certificate; it is a diagnostic
only.

## 8  Intrinsic-prior posterior identifiability

With an intrinsic precision, the conditional posterior precision

\[
P_j=\lambda_jQ_K+D_j
\]

is positive definite if and only if

\[
\operatorname{null}(Q_K)\cap\operatorname{null}(D_j)=\{0\}.
\]

Positive position mass across the relevant components ordinarily makes $D_j$
identify the random-walk null space, so `ridge = 0` is valid.  The
estimated-noise mass floor further prevents floating-point empty positions
from creating a degenerate solve.  Under known noise, positive supplied
precisions and positive position responsibilities play the analogous role.

If the likelihood genuinely does not identify a null-space direction, the
posterior remains singular and fitting stops with an informative error.  An
explicit positive ridge changes the prior to a proper full-rank model; it is a
modeling choice rather than an automatically inserted repair.

## 9  Stored `cavi` state

The derivation maps to the fitted object as follows:

| Mathematical quantity | Fitted-object field |
|---|---|
| $\pi$ | `params$pi` |
| $\sigma_j^2$ | `params$sigma2[j]`, or `NULL` with known `S` |
| supplied $S$ | `measurement_sd` |
| $r_{ik}$ | `gamma[i, k]` |
| $\mu_{jk}$ | `posterior$mean[j, k]` |
| $V_j$ | `posterior$cov[[j]]` |
| $V_{j,kk}$ | `posterior$var[j, k]` or `posterior$diag[j, k]` |
| $\lambda_j$ | `lambda_vec[j]` |
| $Q_K$ | `Q_K` |
| objective sequence | `elbo_trace` |
| plug-in likelihood diagnostic | `loglik_trace` |

`control$prior_proper`, `control$prior_rank`, and `control$prior_logdet`
record whether the trajectory precision was proper and which determinant
convention entered the objective.  `control$noise_model` records the selected
noise mode.

## 10  Relationship to the fixed-$M$ structural model

The multi-ordering structural family is

\[
q(C)\prod_{j=1}^dq(Z_j)q(U_j\mid Z_j).
\]

When $M=1$, $q(Z_j=1)=1$, assignment entropy vanishes, and each conditional
factor $q(U_j\mid Z_j=1)$ becomes the $q_j(U_j)$ derived here.  The structural
expected residual is

\[
E_{jm}=\sum_{i,k}r_{imk}
\{(x_{ij}-\mu_{jmk})^2+V_{jm,kk}\},
\]

and its shared-variance update

\[
\sigma_j^2=\frac1n\sum_{m=1}^Mw_{jm}E_{jm}
\]

reduces to the single-ordering residual update in Section 6.2.  Likewise, the
structural conditional Gaussian, position, lambda, and known-noise formulas
reduce exactly to Sections 4--6.

The important distinction for $M\geq2$ is that the conditional trajectory is
retained as dependent on $Z_j$.  The fixed-$M$ derivation should therefore be
used for partition fits rather than extrapolating a collection of independent
single-ordering `cavi` objects.

## 11  Summary

The current single-ordering algorithm is defined by

\[
q(C,U)=\prod_iq_i(C_i)\prod_jq_j(U_j),
\]

\[
P_j=\lambda_jQ_K+D_j,
\qquad
P_j\mu_j=b_j,
\]

\[
\log r_{ik}
\stackrel{+c}{=}
\log\pi_k
+\sum_j\mathbb E_{q_j}[\log p(x_{ij}\mid C_i=k,U_j)],
\]

and, for estimated noise,

\[
\sigma_j^2
=\frac1n\sum_{i,k}r_{ik}
\{(x_{ij}-\mu_{jk})^2+V_{j,kk}\}.
\]

The smoothness update is $r_Q/A_j$ without a lambda prior and a bounded
one-dimensional optimum when the induced exponential prior is active.  The
recorded `elbo_trace` is a proper ELBO only for a proper $Q_K$; with the
default intrinsic prior it is a fixed-convention pseudo-ELBO.  This distinction
does not change the coordinate formulas, but it is essential when interpreting
the objective as evidence or comparing statistical models.
