---
title: "Known Measurement Noise in Single-Ordering CAVI and Fixed-M Structural CAVI"
output:
  rmarkdown::html_vignette:
    toc: true
    mathjax: default
vignette: >
  %\VignetteIndexEntry{Known Measurement Noise in Single-Ordering CAVI and Fixed-M Structural CAVI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load}
library(MPCurver)
```

This note derives CAVI when measurement standard deviations are known.  It
covers both the ordinary single-ordering model and the current fixed-$M$
structural partition model

\[
q(C)\prod_{j=1}^d q(Z_j)q(U_j\mid Z_j).
\]

The notation matches
`partition_coherent_elbo_derivation.Rmd`.  In particular,

\[
r_{imk}=q(C_i^{(m)}=k),
\qquad
w_{jm}=q(Z_j=m),
\]

and

\[
q(U_j\mid Z_j=m)=N(\mu_{jm},V_{jm}).
\]

The central conclusion is simple but important:

> Supplying `S` changes the likelihood sufficient statistics and removes the
> variance update.  It does not change the structural dependence between
> $Z_j$ and $U_j$, and it does not place assignment weights inside a
> conditional trajectory's Gaussian system.

## 1  Noise specifications accepted by the model

`S` contains standard deviations.  The three supported modes are:

| `S` | Statistical model | Stored noise label |
|---|---|---|
| `NULL` | estimate one $\sigma_j^2$ for feature $j$, shared across orderings | `estimated_feature_variance` |
| length-$d$ vector | known $s_j$, shared across observations and orderings | `known_feature_sd` |
| $n\times d$ matrix | known $s_{ij}$ for every observation and feature | `known_observation_sd` |

For known noise, write

\[
\tau_{ij}=s_{ij}^{-2}.
\]

The observation model is

\[
x_{ij}\mid C_i=k,U_j,S
\sim N(U_{jk},s_{ij}^2)
\]

for a single ordering, and

\[
x_{ij}\mid Z_j=m,C_i^{(m)}=k,U_j,S
\sim N(U_{jk},s_{ij}^2)
\]

for the structural partition model.  The same supplied $s_{ij}$ is used under
every possible ordering because it describes measurement error for the
observed entry $x_{ij}$, not residual lack of fit to a particular ordering.

Supplied standard deviations must be finite and nonnegative.  Internally they
are bounded below by a numerical standard-deviation floor before inversion.
This protects arithmetic involving $\tau_{ij}$; it is unrelated to the
trajectory precision or to random-walk regularization.

## 2  Single-ordering CAVI with known measurement noise

### 2.1  Model and variational family

For one ordering,

\[
C_i\mid\pi\sim\operatorname{Categorical}(\pi),
\]

\[
p(U_j\mid\lambda_j)
\propto
\lambda_j^{r_Q/2}|Q_K|_+^{1/2}
\exp\left(-\frac{\lambda_j}{2}U_j^\top Q_KU_j\right),
\]

where $r_Q=\operatorname{rank}(Q_K)$ and $|Q_K|_+$ is the
pseudo-determinant.  The mean-field family is

\[
q(C,U)=\prod_{i=1}^nq_i(C_i)\prod_{j=1}^dq_j(U_j),
\]

with

\[
r_{ik}=q(C_i=k),
\qquad
q_j(U_j)=N(\mu_j,V_j).
\]

Define

\[
A_j
=\mu_j^\top Q_K\mu_j+\operatorname{tr}(Q_KV_j).
\]

### 2.2  Complete expected likelihood

The expected observation log likelihood is

\[
\mathcal D^S
=-\frac12\sum_{i=1}^n\sum_{k=1}^K\sum_{j=1}^d
r_{ik}
\left[
\log(2\pi s_{ij}^2)
+\tau_{ij}\left\{(x_{ij}-\mu_{jk})^2+V_{j,kk}\right\}
\right].
\]

The $V_{j,kk}$ term is the posterior uncertainty correction

\[
\mathbb E_{q_j}\{(x_{ij}-U_{jk})^2\}
=(x_{ij}-\mu_{jk})^2+V_{j,kk}.
\]

Even though $\log(2\pi s_{ij}^2)$ is constant in $k$, retaining it is
necessary when evaluating the full ELBO.

### 2.3  Full single-ordering objective

Let

\[
\mathcal C
=\sum_{i,k}r_{ik}\log\pi_k
-\sum_{i,k}r_{ik}\log r_{ik}.
\]

For feature $j$, the expected intrinsic-prior term is

\[
\mathcal P_j
=\frac12\left\{
r_Q(\log\lambda_j-\log(2\pi))
+\log|Q_K|_+
\right\}
-\frac{\lambda_j}{2}A_j,
\]

and its Gaussian entropy is

\[
\mathcal H_j
=\frac12\{K\log(2\pi e)+\log|V_j|\}.
\]

If `lambda_sd_prior_rate` is a positive value $\rho$, add

\[
g_\rho(\lambda_j)
=\log(\rho/2)-\frac32\log\lambda_j
-\rho\lambda_j^{-1/2};
\]

otherwise set $g_\rho=0$.  The objective is

\[
\mathcal L_1^S
=\mathcal C+\mathcal D^S
+\sum_{j=1}^d\{\mathcal P_j+g_\rho(\lambda_j)+\mathcal H_j\}.
\]

With an intrinsic $Q_K$, this is a pseudo-ELBO under the package's fixed
pseudo-determinant convention.  With a full-rank precision it is the ordinary
ELBO.

### 2.4  Gaussian update

For each feature, define

\[
D_j
=\operatorname{diag}(d_{j1},\ldots,d_{jK}),
\qquad
d_{jk}=\sum_{i=1}^nr_{ik}\tau_{ij},
\]

and

\[
b_{jk}=\sum_{i=1}^nr_{ik}\tau_{ij}x_{ij}.
\]

Completing the square yields

\[
P_j=\lambda_jQ_K+D_j,
\qquad
P_j\mu_j=b_j,
\qquad
V_j=P_j^{-1}.
\]

For feature-level known noise, $s_{ij}=s_j$ and therefore

\[
d_{jk}=N_k/s_j^2,
\qquad
b_{jk}=s_j^{-2}\sum_i r_{ik}x_{ij},
\qquad
N_k=\sum_i r_{ik}.
\]

For observation-level known noise, two observations assigned to the same
position contribute different precision according to their supplied
measurement uncertainty.

### 2.5  Position-responsibility update

For each observation,

\[
\log r_{ik}
\stackrel{+c}{=}
\log\pi_k
-\frac12\sum_{j=1}^d
\left[
\log(2\pi s_{ij}^2)
+\tau_{ij}\{(x_{ij}-\mu_{jk})^2+V_{j,kk}\}
\right].
\]

The log-normalizing term is independent of $k$ and therefore cancels in the
softmax, but including it in a full log-density matrix is algebraically
harmless.  The uncertainty correction remains essential.

### 2.6  Position prior and lambda

For an adaptive position prior,

\[
\pi_k=\frac1n\sum_i r_{ik};
\]

a fixed prior is left unchanged.  The lambda update uses

\[
A_j=\mu_j^\top Q_K\mu_j+\operatorname{tr}(Q_KV_j).
\]

Without the optional scale prior,

\[
\lambda_j=r_Q/A_j.
\]

With $\tau_j=\lambda_j^{-1/2}\sim\operatorname{Exp}(\rho)$, the implementation
maximizes

\[
\frac{r_Q-3}{2}\log\lambda_j
-\frac{A_j}{2}\lambda_j
-\rho\lambda_j^{-1/2}
\]

within the supplied lambda bounds.

### 2.7  There is no variance update

When `S` is non-`NULL`, $s_{ij}^2$ is fixed data supplied by the user.  The
fit therefore stores `params$sigma2 = NULL`, skips every estimated-variance
update, and preserves the original vector or matrix specification in
`measurement_sd`.

## 3  Fixed-$M$ structural partition with known noise

### 3.1  Structural factorization

For $M\geq2$, the variational family is

\[
q(C,Z,U)
=\left\{\prod_{m=1}^M\prod_{i=1}^nq_{im}(C_i^{(m)})\right\}
\left\{\prod_{j=1}^dq_j(Z_j)q_j(U_j\mid Z_j)\right\}.
\]

Write

\[
q(U_j\mid Z_j=m)=N(\mu_{jm},V_{jm}).
\]

These $M$ Gaussians are conditional alternatives for one trajectory, not $M$
independent posterior trajectories that coexist in the generative state.

### 3.2  Conditional Gaussian branches

For ordering $m$, define

\[
d_{jmk}=\sum_{i=1}^nr_{imk}\tau_{ij},
\qquad
b_{jmk}=\sum_{i=1}^nr_{imk}\tau_{ij}x_{ij},
\]

\[
D_{jm}=\operatorname{diag}(d_{jm1},\ldots,d_{jmK}).
\]

The structural conditional update is

\[
\boxed{
P_{jm}=\lambda_{jm}Q_K+D_{jm},
\qquad
P_{jm}\mu_{jm}=b_{jm},
\qquad
V_{jm}=P_{jm}^{-1}.
}
\]

The assignment probability $w_{jm}$ is absent from this Gaussian system.  For
$w_{jm}>0$, it multiplies the complete local conditional block and cancels
from the coordinate optimum.  If a stored weight is exactly zero, the
objective does not uniquely determine that zero-probability conditional
factor; the implementation uses the same continuous branch update so that
the branch remains evaluable in the next assignment step.

### 3.3  Ordering-specific position responsibilities

Feature assignments do affect the ordering-position update.  For each
$m,i,k$,

\[
\boxed{
\log r_{imk}
\stackrel{+c}{=}
\log\pi_k^{(m)}
-\frac12\sum_{j=1}^dw_{jm}
\left[
\log(2\pi s_{ij}^2)
+\tau_{ij}\{(x_{ij}-\mu_{jmk})^2+V_{jm,kk}\}
\right].
}
\]

Thus a feature contributes strongly to the position posterior of an ordering
only when $q(Z_j=m)$ is large.  This is the correct location for $w_{jm}$ in
the position coordinate.

For adaptive position priors,

\[
\pi_k^{(m)}=\frac1n\sum_i r_{imk};
\]

fixed position priors remain unchanged.

### 3.4  Complete known-noise local block

The data contribution for feature $j$ and ordering $m$ is

\[
\mathcal D_{jm}^S
=-\frac12\sum_{i=1}^n\sum_{k=1}^Kr_{imk}
\left[
\log(2\pi s_{ij}^2)
+\tau_{ij}\{(x_{ij}-\mu_{jmk})^2+V_{jm,kk}\}
\right].
\]

Set

\[
A_{jm}
=\mu_{jm}^\top Q_K\mu_{jm}
+\operatorname{tr}(Q_KV_{jm}).
\]

The conditional trajectory-prior contribution is

\[
\mathcal P_{jm}
=\frac12\left\{
r_Q(\log\lambda_{jm}-\log(2\pi))
+\log|Q_K|_+
\right\}
-\frac{\lambda_{jm}}2A_{jm},
\]

and the entropy is

\[
\mathcal H_{jm}
=\frac12\{K\log(2\pi e)+\log|V_{jm}|\}.
\]

The exact local score stored by the structural backend is

\[
\boxed{
B_{jm}^S
=\mathcal D_{jm}^S
+\mathcal P_{jm}
+g_\rho(\lambda_{jm})
+\mathcal H_{jm}.
}
\]

All four pieces are conditional on the same feature assignment and enter the
total objective multiplied by $w_{jm}$.

### 3.5  Assignment update

Let $a_m=\log\omega_m$ for a fixed or adaptive point ordering prior, and let
$a_m=\mathbb E[\log\omega_m]$ for the deprecated Dirichlet compatibility
mode.  At temperature $T$,

\[
\boxed{
w_{jm}
=\frac{\exp\{(B_{jm}^S+a_m)/T\}}
{\sum_{\ell=1}^M\exp\{(B_{j\ell}^S+a_\ell)/T\}}.
}
\]

Known measurement noise changes $B_{jm}$ through $s_{ij}^{-2}$ and the
observation-specific log constants.  It does not otherwise alter the
assignment softmax.

For an adaptive point ordering prior,

\[
\omega_m=d^{-1}\sum_jw_{jm}.
\]

For a fixed prior, $\omega$ is unchanged.  In the Dirichlet compatibility
mode with symmetric prior concentration $\alpha_0$,

\[
q(\omega)
=\operatorname{Dirichlet}\left(
\alpha_0+\sum_jw_{j1},\ldots,
\alpha_0+\sum_jw_{jM}
\right).
\]

### 3.6  Total fixed-$M$ objective

For each ordering, define the cell block

\[
\mathcal C_m
=\sum_{i,k}r_{imk}\log\pi_k^{(m)}
-\sum_{i,k}r_{imk}\log r_{imk}.
\]

Let

\[
H_Z=-\sum_{j,m}w_{jm}\log w_{jm}.
\]

For a point ordering prior,

\[
\mathcal A_T
=\sum_{j,m}w_{jm}\log\omega_m+T H_Z.
\]

For the Dirichlet compatibility mode,

\[
\mathcal A_T
=\sum_{j,m}w_{jm}\mathbb E[\log\omega_m]
+T H_Z
-\operatorname{KL}\{q(\omega)\|p(\omega)\}.
\]

The known-noise structural objective is

\[
\boxed{
\mathcal L_T^S
=\sum_{m=1}^M\mathcal C_m
+\sum_{j=1}^d\sum_{m=1}^Mw_{jm}B_{jm}^S
+\mathcal A_T.
}
\]

At $T=1$, this is the proper-prior ELBO or intrinsic-prior pseudo-ELBO.  At
$T>1$, it is an entropy-tempered objective.

### 3.7  Ordering-specific lambda update

The same second moment $A_{jm}$ used in the local prior produces

\[
\lambda_{jm}=r_Q/A_{jm}
\]

when the optional scale prior is absent.  With rate $\rho>0$, the bounded
branchwise objective is

\[
\frac{r_Q-3}{2}\log\lambda_{jm}
-\frac{A_{jm}}2\lambda_{jm}
-\rho\lambda_{jm}^{-1/2}.
\]

This update is independent of whether $S$ is a vector or matrix; the noise
specification has already affected $\mu_{jm}$ and $V_{jm}$ through $D_{jm}$
and $b_{jm}$.

## 4  Estimated noise versus known noise

The current structural model has one estimated variance $\sigma_j^2$ per
feature, shared across all orderings.  It never estimates a separate residual
variance for each conditional branch.

For reference, define the structural expected residual

\[
E_{jm}
=\sum_{i,k}r_{imk}
\{(x_{ij}-\mu_{jmk})^2+V_{jm,kk}\}.
\]

When `S = NULL`,

\[
\boxed{
\sigma_j^2=\frac1n\sum_{m=1}^Mw_{jm}E_{jm}.
}
\]

When `S` is supplied, this coordinate is removed rather than replaced by an
ordering-specific parameter.

| Component | Estimated noise (`S = NULL`) | Known noise (`S` supplied) |
|---|---|---|
| Variance | one fitted $\sigma_j^2$ per feature, shared across $m$ | fixed $s_j^2$ or $s_{ij}^2$ |
| Gaussian diagonal | $N_{mk}/\sigma_j^2$ | $\sum_i r_{imk}/s_{ij}^2$ |
| Gaussian right-hand side | $\sum_i r_{imk}x_{ij}/\sigma_j^2$ | $\sum_i r_{imk}x_{ij}/s_{ij}^2$ |
| Position likelihood | weighted by $w_{jm}$ | weighted by $w_{jm}$ |
| Conditional Gaussian system | no assignment weight | no assignment weight |
| Variance update | shared update above | none |
| Conditional prior and entropy | unchanged | unchanged |
| Assignment softmax | complete estimated-noise $B_{jm}$ | complete known-noise $B_{jm}^S$ |

Known `S` does not remove the model's ability to compare orderings.  Ordering
evidence still differs through the fitted conditional mean, posterior
uncertainty, expected residual under $r_{imk}$, smoothness cost, Gaussian
entropy, and position fit.  What disappears is only adaptation of an unknown
residual variance.

## 5  Feature-level and observation-level special cases

### 5.1  Feature-level known standard deviation

If $s_{ij}=s_j$, then

\[
D_{jm}
=\operatorname{diag}(N_{m1}/s_j^2,\ldots,N_{mK}/s_j^2),
\]

and

\[
b_{jm,k}=s_j^{-2}\sum_i r_{imk}x_{ij}.
\]

The likelihood normalizer for feature $j$ is simply
$-n\log(2\pi s_j^2)/2$ in every conditional branch.  It is constant across
orderings for that feature but remains part of the full fixed-$M$ objective.

### 5.2  Observation-level known standard deviation

If $s_{ij}$ varies over observations, the precision-weighted effective mass

\[
d_{jmk}=\sum_i r_{imk}s_{ij}^{-2}
\]

need not be proportional to $N_{mk}$.  An observation with a smaller supplied
standard deviation contributes more strongly to both the posterior precision
and posterior mean.  The exact log normalizer

\[
-\frac12\sum_i\log(2\pi s_{ij}^2)
\]

must be retained when evaluating the local block, although it is constant in
$k$ during a position softmax.

## 6  Intrinsic-prior identifiability under known noise

For an intrinsic RW$(q)$ precision, $Q_K$ has a $q$-dimensional null space.
The conditional posterior is proper if and only if

\[
\operatorname{null}(Q_K)\cap\operatorname{null}(D_{jm})=\{0\}.
\]

Because valid known-noise precisions are strictly positive after numerical
flooring, positive position mass in every component makes $D_{jm}$ positive
diagonal and identifies all null-space directions.  The implementation's
softmax stabilization prevents finite position logits from underflowing to
exactly empty components.

The feature-assignment probability does not enter $D_{jm}$.  Therefore a
small or zero $w_{jm}$ does not itself destroy conditional posterior
identifiability.  Nevertheless, if externally supplied or genuinely
degenerate position responsibilities fail to identify the intrinsic null
space, the posterior precision remains singular.  The backend reports that
identifiability failure instead of adding automatic jitter.

An explicitly supplied ridge makes $Q_K$ full rank and changes the prior to a
proper Gaussian.  It is a modeling choice, not a requirement caused by a
feature being unlikely to use an ordering.

## 7  Annealing and fixed-$M$ interpretation

During deterministic annealing, only the feature-assignment entropy is
multiplied by $T$.  Consequently objective values from different temperatures
represent different optimization criteria and are not directly comparable.
The exact structural objective is evaluated at $T=1$, and monotonicity is
assessed only within the subsequent fixed-$T=1$ sweeps.  Continuation through
`do_mpcurve()` also uses $T=1$.

The known-noise objective is a diagnostic for one prespecified $M$.  It is not
a cross-$M$ selection criterion.  In particular, an adaptive ordering prior
can give an additional nearly unused ordering negligible feature mass without
adding a calibrated complexity penalty, and an intrinsic prior yields a
pseudo-ELBO rather than a normalized model evidence.  The package therefore
does not currently perform greedy forward or backward dimension selection.

## 8  Canonical fitted state

For a fixed-$M$ known-noise structural fit:

- `measurement_sd` contains the supplied length-$d$ vector or $n\times d$
  matrix;
- `params$sigma2` is `NULL`;
- `gamma[[m]]` contains $r_{imk}$;
- `position_pi[[m]]` contains $\pi_k^{(m)}$;
- `partition$pi_weights` on the normalized `mpcurve` object, and
  `pi_weights` on the raw structural fit, contain $w_{jm}$;
- `conditional_posterior$mean[[m]]` and `$cov[[m]]` contain $\mu_{jm}$ and
  $V_{jm}$;
- `lambda_mat[j, m]` contains $\lambda_{jm}$; and
- `local_blocks[j, m]` contains $B_{jm}^S$.

This top-level state is the authority for both plotting and continuation.
Per-ordering `$fits` entries are read-only compatibility views derived from
it.

## 9  Summary

For single-ordering and fixed-$M$ structural CAVI alike, known measurement
noise changes the sufficient statistics to

\[
d_{jmk}=\sum_i r_{imk}s_{ij}^{-2},
\qquad
b_{jmk}=\sum_i r_{imk}s_{ij}^{-2}x_{ij}.
\]

The conditional Gaussian update remains

\[
P_{jm}=\lambda_{jm}Q_K+D_{jm},
\qquad
P_{jm}\mu_{jm}=b_{jm}.
\]

In the partition model, assignment weights enter the ordering-position update,
the complete-local-block average, and the assignment-prior updates.  They do
not rescale a conditional branch's posterior precision or right-hand side.
There is no estimated variance update when `S` is supplied; when noise is
estimated instead, the one $\sigma_j^2$ for feature $j$ is shared across all
orderings.

These identities make the known-noise and estimated-noise implementations two
likelihood variants of the same structural variational model rather than two
different partition factorizations.
