---
title: "Fixed-M Structural Variational Inference for the Multi-Ordering MPCurve Model"
output:
  rmarkdown::html_vignette:
    toc: true
    mathjax: default
vignette: >
  %\VignetteIndexEntry{Fixed-M Structural Variational Inference for the Multi-Ordering MPCurve Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load}
library(MPCurver)
```

This note derives the model and coordinate updates implemented for a fixed
number of orderings, $M \geq 2$.  The implementation uses the structural
variational family

\[
q(C)\prod_{j=1}^d q(Z_j)q(U_j\mid Z_j),
\]

not a collection of independent posterior trajectories for every
feature--ordering pair.  The $M$ Gaussian objects stored for feature $j$ are
the mutually exclusive conditional branches
$q(U_j\mid Z_j=m)$ of one latent trajectory.

The derivation is deliberately implementation-specific.  It covers:

- the single canonical structural state used by fitting and continuation;
- estimated feature variance shared across all orderings;
- known feature-level or observation-level measurement standard deviations;
- intrinsic random-walk priors and their pseudo-determinants;
- the complete conditional local block used to update $q(Z_j)$;
- fixed and adaptive ordering-usage priors, plus the deprecated Dirichlet
  compatibility path;
- deterministic annealing and the exact $T=1$ objective; and
- why the recorded fixed-$M$ objective is not used to select $M$.

Throughout, $M$ is fixed before fitting.  The current implementation does not
perform forward or backward selection of $M$.

## 1  Indices, variables, and canonical quantities

Let

- $i=1,\ldots,n$ index observations;
- $j=1,\ldots,d$ index features;
- $m=1,\ldots,M$ index candidate orderings; and
- $k=1,\ldots,K$ index discretized positions along an ordering.

The observed data matrix is $X=(x_{ij})\in\mathbb R^{n\times d}$.  The latent
variables are:

\[
C_i^{(m)}\in\{1,\ldots,K\},\qquad
Z_j\in\{1,\ldots,M\},\qquad
U_j\in\mathbb R^K.
\]

$C_i^{(m)}$ is observation $i$'s position under ordering $m$.  $Z_j$ chooses
which ordering explains feature $j$.  Conditional on $Z_j=m$, the vector
$U_j=(U_{j1},\ldots,U_{jK})^\top$ is the trajectory of feature $j$ along
ordering $m$.

The variational parameters used below are

\[
r_{imk}=q(C_i^{(m)}=k),\qquad
w_{jm}=q(Z_j=m),
\]

and

\[
q(U_j\mid Z_j=m)
=q_{jm}(U_j)
=N(\mu_{jm},V_{jm}).
\]

For every $i,m$ and $j$,

\[
\sum_{k=1}^K r_{imk}=1,
\qquad
\sum_{m=1}^M w_{jm}=1.
\]

Define the position mass and the trajectory second moment

\[
N_{mk}=\sum_{i=1}^n r_{imk},
\]

\[
A_{jm}
:=
\mathbb E_{q_{jm}}[U_j^\top Q_KU_j]
=
\mu_{jm}^\top Q_K\mu_{jm}
+\operatorname{tr}(Q_KV_{jm}).
\]

The raw structural fit stores these quantities once, at top level:

- `gamma[[m]]` stores $(r_{imk})_{i,k}$;
- `position_pi[[m]]` stores $\pi^{(m)}$;
- `pi_weights[j, m]` stores $w_{jm}$;
- `conditional_posterior$mean[[m]][j, ]` stores $\mu_{jm}$;
- `conditional_posterior$cov[[m]][[j]]` stores $V_{jm}$;
- `lambda_mat[j, m]` stores $\lambda_{jm}$; and
- `params$sigma2[j]` stores the one variance $\sigma_j^2$ shared by all
  orderings when measurement noise is estimated.

The per-ordering objects in `$fits` are derived compatibility views.  They are
not independent variational states and are not read back during continuation.

## 2  Probabilistic model

### 2.1  Position variables

For each ordering $m$,

\[
C_i^{(m)}\mid\pi^{(m)}
\sim\operatorname{Categorical}(\pi^{(m)}),
\qquad
\sum_{k=1}^K\pi_k^{(m)}=1.
\]

The implementation either holds $\pi^{(m)}$ fixed or estimates it as a point
parameter.

### 2.2  Feature assignment

For every feature,

\[
Z_j\mid\omega\sim\operatorname{Categorical}(\omega),
\qquad
\sum_{m=1}^M\omega_m=1.
\]

The main interfaces support two point-parameter conventions:

1. **fixed**: $\omega$ is supplied, or is uniform if it is not supplied;
2. **adaptive**: $\omega$ is profiled from the current expected feature
   usage.

A deprecated compatibility path instead assigns a symmetric Dirichlet prior
to $\omega$ and maintains $q(\omega)$.  Section 6.7 gives all three objective
forms.

### 2.3  Conditional trajectory prior

Let $Q_K$ be the unit-scale random-walk precision and let

\[
r_Q=\operatorname{rank}(Q_K),
\qquad
\log|Q_K|_+
=\sum_{\nu_\ell>0}\log\nu_\ell,
\]

where $|\cdot|_+$ is the pseudo-determinant.  Conditional on $Z_j=m$,

\[
p(U_j\mid Z_j=m,\lambda_{jm})
\propto
\lambda_{jm}^{r_Q/2}|Q_K|_+^{1/2}
\exp\left\{-\frac{\lambda_{jm}}2U_j^\top Q_KU_j\right\}.
\]

With an intrinsic RW$(q)$ prior and no ridge, $r_Q=K-q$.  This is an
intrinsic Gaussian density defined only up to its null-space reference
measure.  Consequently the recorded quantity is a pseudo-ELBO under a fixed
pseudo-determinant convention.  If $Q_K$ is made full rank explicitly, the
same expression becomes the ordinary proper Gaussian log density with
$r_Q=K$ and the usual determinant.

Optionally, the implementation places an exponential prior on the trajectory
standard-deviation scale

\[
\tau_{jm}=\lambda_{jm}^{-1/2},
\qquad
\tau_{jm}\sim\operatorname{Exp}(\rho),
\]

which induces

\[
g_\rho(\lambda_{jm})
=\log p(\lambda_{jm})
=\log(\rho/2)-\frac32\log\lambda_{jm}
-\rho\lambda_{jm}^{-1/2}.
\]

When `lambda_sd_prior_rate = NULL`, $g_\rho$ is defined to be zero.

### 2.4  Observation model with estimated noise

When `S = NULL`, each feature has one variance shared across observations,
positions, and orderings:

\[
x_{ij}\mid Z_j=m,C_i^{(m)}=k,U_j,\sigma_j^2
\sim N(U_{jk},\sigma_j^2).
\]

The estimated-noise parameterization contains exactly one $\sigma_j^2$ for
feature $j$.  That shared parameterization is essential to the current model
and to the update in Section 6.5.

### 2.5  Observation model with known noise

When `S` is supplied, it contains standard deviations, not variances.  A
length-$d$ vector means $s_{ij}=s_j$ for every observation, while an
$n\times d$ matrix supplies $s_{ij}$ separately.  In either case,

\[
x_{ij}\mid Z_j=m,C_i^{(m)}=k,U_j,S
\sim N(U_{jk},s_{ij}^2).
\]

The implementation replaces zero or extremely small supplied standard
deviations by a numerical floor before constructing $s_{ij}^{-2}$.  No
$\sigma_j^2$ parameter exists in this mode.

## 3  Structural variational family

Write

\[
q(C)=\prod_{m=1}^M\prod_{i=1}^n q_{im}(C_i^{(m)}).
\]

Ignoring $q(\omega)$ for the moment, the complete structural family is

\[
q(C,Z,U)
=
\left\{\prod_{m=1}^M\prod_{i=1}^nq_{im}(C_i^{(m)})\right\}
\left\{\prod_{j=1}^d q_j(Z_j)q_j(U_j\mid Z_j)\right\}.
\]

The dependence between assignment and trajectory is retained:

\[
q_j(U_j\mid Z_j=m)=q_{jm}(U_j).
\]

This matters both mathematically and computationally.  The entropy of one
feature's structural factor is

\[
H\{q_j(Z_j,U_j)\}
=H\{q_j(Z_j)\}
+\sum_{m=1}^Mw_{jm}H(q_{jm}),
\]

so the conditional Gaussian entropy is weighted by the probability of the
corresponding assignment branch.  Likelihood, trajectory prior, optional
lambda prior, and Gaussian entropy therefore enter one complete local block.

## 4  The fixed-$M$ structural objective

### 4.1  Cell-position block

For ordering $m$, define

\[
\mathcal C_m
=
\sum_{i=1}^n\sum_{k=1}^K
r_{imk}\log\pi_k^{(m)}
-
\sum_{i=1}^n\sum_{k=1}^K
r_{imk}\log r_{imk}.
\]

This is the expected position-prior log density plus the entropy of
$q(C^{(m)})$.

### 4.2  Complete conditional local block

For feature $j$ and ordering $m$, define

\[
B_{jm}
=\mathcal D_{jm}
+\mathcal P_{jm}
+g_\rho(\lambda_{jm})
+\mathcal H_{jm},
\]

where $\mathcal D_{jm}$ is the expected observation log likelihood,
$\mathcal P_{jm}$ is the expected conditional trajectory log prior, and
$\mathcal H_{jm}=H(q_{jm})$.

For estimated noise,

\[
\mathcal D_{jm}^{\mathrm{est}}
=-\frac12\sum_{i=1}^n\sum_{k=1}^K r_{imk}
\left[
\log(2\pi\sigma_j^2)
+\frac{(x_{ij}-\mu_{jmk})^2+V_{jm,kk}}{\sigma_j^2}
\right].
\]

For known noise,

\[
\mathcal D_{jm}^{S}
=-\frac12\sum_{i=1}^n\sum_{k=1}^K r_{imk}
\left[
\log(2\pi s_{ij}^2)
+\frac{(x_{ij}-\mu_{jmk})^2+V_{jm,kk}}{s_{ij}^2}
\right].
\]

The prior term is

\[
\mathcal P_{jm}
=\frac12\left\{
r_Q\big(\log\lambda_{jm}-\log(2\pi)\big)
+\log|Q_K|_+
\right\}
-\frac{\lambda_{jm}}2A_{jm},
\]

and the Gaussian entropy is

\[
\mathcal H_{jm}
=\frac12\left\{
K\log(2\pi e)+\log|V_{jm}|
\right\}.
\]

Thus $B_{jm}$ is complete: no Gaussian entropy or trajectory-prior term is
added elsewhere without the same assignment weight.

### 4.3  Assignment block

Let

\[
H_Z=-\sum_{j=1}^d\sum_{m=1}^M w_{jm}\log w_{jm}.
\]

For fixed or adaptive point $\omega$, the temperature-$T$ assignment block is

\[
\mathcal A_T
=\sum_{j,m}w_{jm}\log\omega_m+T H_Z.
\]

For the Dirichlet compatibility mode,

\[
\mathcal A_T
=\sum_{j,m}w_{jm}\mathbb E_{q(\omega)}[\log\omega_m]
+T H_Z
-\operatorname{KL}\{q(\omega)\|p(\omega)\}.
\]

### 4.4  Total objective

The quantity evaluated by the structural backend is

\[
\boxed{
\mathcal L_T
=\sum_{m=1}^M\mathcal C_m
+\sum_{j=1}^d\sum_{m=1}^M w_{jm}B_{jm}
+\mathcal A_T.
}
\]

At $T=1$, this is the structural ELBO for proper trajectory priors and the
fixed-convention structural pseudo-ELBO for an intrinsic prior.  For $T>1$,
it is an entropy-tempered optimization objective, not the ordinary ELBO.

## 5  Sufficient statistics for the conditional Gaussian branches

### 5.1  Estimated shared variance

For a fixed ordering $m$, define

\[
D_{jm}
=\operatorname{diag}\left(
\frac{N_{m1}}{\sigma_j^2},\ldots,
\frac{N_{mK}}{\sigma_j^2}
\right),
\]

and

\[
b_{jm,k}
=\frac{1}{\sigma_j^2}
\sum_{i=1}^n r_{imk}x_{ij}.
\]

### 5.2  Known measurement standard deviations

Set $\tau_{ij}=s_{ij}^{-2}$.  Then

\[
D_{jm}
=\operatorname{diag}\left(
\sum_i r_{im1}\tau_{ij},\ldots,
\sum_i r_{imK}\tau_{ij}
\right),
\]

\[
b_{jm,k}
=\sum_{i=1}^n r_{imk}\tau_{ij}x_{ij}.
\]

If `S` is a length-$d$ vector, $\tau_{ij}=s_j^{-2}$ and these expressions
reduce to the estimated-noise sufficient statistics with $\sigma_j^2$
formally replaced by the fixed value $s_j^2$.

## 6  Coordinate updates

### 6.1  Conditional trajectory update

For $w_{jm}>0$, all terms involving $q_{jm}$ occur inside
$w_{jm}B_{jm}$.  Dividing the coordinate objective by this positive scalar
shows that the assignment probability does not alter the conditional
posterior.  Completing the square gives

\[
P_{jm}=\lambda_{jm}Q_K+D_{jm},
\qquad
P_{jm}\mu_{jm}=b_{jm},
\qquad
V_{jm}=P_{jm}^{-1}.
\]

Both the likelihood precision $D_{jm}$ and right-hand side $b_{jm}$ use the
full conditional data contribution.  Assignment probabilities do not appear
inside either quantity.

If $w_{jm}=0$ exactly, the ELBO is indifferent to the value of the
zero-probability conditional factor.  The implementation uses the continuous
extension of the positive-weight update above and keeps a well-defined
conditional branch available for the next assignment update.  This convention
is also what makes the local score $B_{jm}$ evaluable if a branch later regains
positive probability.

### 6.2  Position-responsibility update

The terms involving $r_{imk}$ are the position prior, position entropy, and
the assignment-weighted expected likelihoods.  For estimated noise,

\[
\log r_{imk}
\stackrel{+c}{=}
\log\pi_k^{(m)}
-\frac12\sum_{j=1}^d w_{jm}
\left[
\log(2\pi\sigma_j^2)
+\frac{(x_{ij}-\mu_{jmk})^2+V_{jm,kk}}{\sigma_j^2}
\right].
\]

For known noise,

\[
\log r_{imk}
\stackrel{+c}{=}
\log\pi_k^{(m)}
-\frac12\sum_{j=1}^d w_{jm}
\left[
\log(2\pi s_{ij}^2)
+\frac{(x_{ij}-\mu_{jmk})^2+V_{jm,kk}}{s_{ij}^2}
\right].
\]

Here $+c$ denotes an additive constant independent of $k$.  The posterior
variance term $V_{jm,kk}$ is required; replacing it by zero would turn the
update into a plug-in approximation.

The implementation evaluates these logits with a row-wise log-sum-exp and
then applies a very small categorical-probability floor to prevent floating
point underflow to exact zero.  This simplex stabilization is not a precision
ridge.

### 6.3  Position-prior update

For an adaptive position prior,

\[
\pi_k^{(m)}
=\frac1n\sum_{i=1}^n r_{imk}.
\]

For a fixed position prior, $\pi^{(m)}$ is left unchanged.

### 6.4  Expected residuals

For subsequent parameter updates, define

\[
E_{jm}
=\sum_{i=1}^n\sum_{k=1}^K r_{imk}
\left\{(x_{ij}-\mu_{jmk})^2+V_{jm,kk}\right\}.
\]

This is the full expected squared residual for feature $j$ under conditional
branch $m$.

### 6.5  Shared estimated-variance update

Because $\sum_mw_{jm}=1$, the $\sigma_j^2$-dependent part of the objective is

\[
-\frac n2\log\sigma_j^2
-\frac{1}{2\sigma_j^2}\sum_{m=1}^M w_{jm}E_{jm}.
\]

Its maximizer is

\[
\boxed{
\sigma_j^2
=\frac1n\sum_{m=1}^M w_{jm}E_{jm}.
}
\]

The implementation clips this update to `[sigma_min, sigma_max]`.  With known
`S`, this coordinate does not exist and `params$sigma2` is `NULL`.

### 6.6  Ordering-specific lambda update

Without the optional scale prior, the $\lambda_{jm}$-dependent part of
$B_{jm}$ is

\[
\frac{r_Q}{2}\log\lambda_{jm}
-\frac{A_{jm}}2\lambda_{jm},
\]

so

\[
\boxed{
\lambda_{jm}=\frac{r_Q}{A_{jm}}.
}
\]

With $\tau_{jm}\sim\operatorname{Exp}(\rho)$, the bounded one-dimensional
objective is

\[
\frac{r_Q-3}{2}\log\lambda_{jm}
-\frac{A_{jm}}2\lambda_{jm}
-\rho\lambda_{jm}^{-1/2}.
\]

It is optimized on the log scale and clipped to
`[lambda_min, lambda_max]`.  Because the whole conditional block is multiplied
by $w_{jm}$, a positive assignment weight scales all its lambda-dependent
terms equally and does not change this branchwise maximizer.  The implementation
uses the same continuous branch update when the stored weight is zero.  If
`fix_lambda = TRUE`, the supplied $\lambda_{jm}$ values are retained.

### 6.7  Assignment and ordering-usage updates

Given the complete local blocks, the assignment coordinate is

\[
\boxed{
w_{jm}
=\frac{\exp\{(B_{jm}+a_m)/T\}}
{\sum_{\ell=1}^M\exp\{(B_{j\ell}+a_\ell)/T\}},
}
\]

where

\[
a_m=
\begin{cases}
\log\omega_m,
&\text{fixed or adaptive point prior},\\
\mathbb E_{q(\omega)}[\log\omega_m],
&\text{Dirichlet compatibility mode}.
\end{cases}
\]

For the adaptive point prior,

\[
\omega_m
=\frac{1}{d}\sum_{j=1}^d w_{jm}.
\]

For a symmetric Dirichlet prior
$p(\omega)=\operatorname{Dirichlet}(\alpha_0,\ldots,\alpha_0)$,

\[
q(\omega)
=\operatorname{Dirichlet}(\alpha_1,\ldots,\alpha_M),
\qquad
\alpha_m=\alpha_0+\sum_jw_{jm},
\]

\[
\mathbb E[\log\omega_m]
=\psi(\alpha_m)-\psi\left(\sum_\ell\alpha_\ell\right).
\]

The fixed and adaptive point-prior modes are the current interfaces.  The
Dirichlet form is retained only to explain the exact objective of the
backward-compatibility arguments.

## 7  Intrinsic priors and posterior identifiability

For an intrinsic RW$(q)$ prior, $Q_K$ is positive semidefinite with a
$q$-dimensional null space.  The conditional posterior precision

\[
P_{jm}=\lambda_{jm}Q_K+D_{jm}
\]

is positive definite precisely when the likelihood identifies every prior
null-space direction:

\[
\operatorname{null}(Q_K)\cap\operatorname{null}(D_{jm})=\{0\}.
\]

This condition is often easy to satisfy.  With finite position logits, every
$r_{imk}$ is theoretically positive; the estimated-noise $D_{jm}$ then has a
positive diagonal.  The same is true under known positive precisions.  The
categorical stabilization used in code also prevents numerical softmax
underflow from creating empty position columns.

Crucially, $D_{jm}$ is built from the conditional likelihood and is present
even if the feature's current assignment probability for ordering $m$ is
zero.  A zero assignment probability therefore does not, by itself, remove
the likelihood information needed to identify the intrinsic null space.

This does not make every intrinsic posterior automatically proper.  If the
position responsibilities and measurement precisions genuinely fail to
identify a null-space direction, $P_{jm}$ remains singular.  The structural
backend then reports an identifiability error.  It does not silently add
precision jitter.  An explicitly requested ridge changes the statistical
prior and is treated as a proper full-rank precision; it is not required merely
to compensate for a small feature-assignment probability.

## 8  Annealing, convergence, and hard assignments

The fitting schedule replaces $H_Z$ by $T H_Z$ with $T$ decreasing from
`T_start` to `T_end`.  At a fixed temperature, the softmax in Section 6.7 is
the exact coordinate maximizer for the corresponding entropy-tempered
objective.  However,

\[
\mathcal L_{T_1}\neq\mathcal L_{T_2}
\qquad\text{when }T_1\neq T_2,
\]

so objective values recorded at different temperatures must not be compared
as if they came from one monotone sequence.

After annealing, the implementation ensures that the state is evaluated at
$T=1$ and performs exact-temperature coordinate sweeps.  Monotonicity and
convergence diagnostics belong to this fixed-$T=1$ segment.  `do_mpcurve()`
continues a structural fit only with $T=1$ sweeps from the top-level canonical
state.

If `hard_assign_final = TRUE`, the final $w_{jm}$ values are projected to
one-hot assignments.  That projection is a reporting or downstream-use
choice, not the soft variational coordinate maximizer, and its final objective
need not improve over the preceding soft state.

## 9  Reduction to one ordering

When $M=1$,

\[
w_{j1}=1,
\qquad
H_Z=0,
\qquad
\omega_1=1.
\]

There is one $q(C)$ factor and one Gaussian branch per feature.  The structural
objective reduces to the ordinary single-ordering CAVI objective, the shared
variance update becomes

\[
\sigma_j^2=E_{j1}/n,
\]

and the conditional Gaussian and lambda updates become the corresponding
single-ordering updates.  This is a mathematical reduction; the partition
backend itself requires $M\geq2$, while `cavi()` is used for $M=1$.

## 10  Why the fixed-$M$ objective does not select $M$

`objective_history` is designed to optimize and diagnose one prespecified
$M$.  It is not exposed as a complexity-penalized model-selection score.

Several facts prevent a naive cross-$M$ comparison:

1. Under an intrinsic prior the recorded quantity uses a fixed
   pseudo-determinant convention rather than a normalized marginal likelihood.
2. With an adaptive ordering-usage prior, an additional nearly empty ordering
   can receive almost no assignment mass without paying a calibrated
   model-complexity cost.
3. Different values of $M$ introduce different position variables, conditional
   branches, and optimization landscapes; local variational optima are not a
   substitute for a common marginal-evidence calculation.
4. Values from different annealing temperatures are not comparable even within
   one fit.

A fixed uniform $\omega$ does contribute an explicit
$-d\log M$ assignment term, but that alone does not turn the current raw
objective into a validated dimension-selection criterion.  For this reason,
forward and backward greedy selection are temporarily unavailable.  Users
should choose $M$ externally or compare fixed-$M$ fits using a separately
defined and validated criterion.

## 11  Exact correspondence with the canonical implementation

One complete structural sweep performs the following conceptual updates:

1. update every conditional Gaussian branch from the current $q(C^{(m)})$;
2. update every $q(C^{(m)})$ using feature likelihoods weighted by $w_{jm}$;
3. update adaptive position priors;
4. update the single shared $\sigma_j^2$ vector when `S = NULL`;
5. update the ordering-specific $\lambda_{jm}$ values unless fixed;
6. recompute the complete $B_{jm}$ matrix;
7. update $q(Z_j)$ and the selected ordering-usage prior; and
8. evaluate $\sum_m\mathcal C_m+\sum_{jm}w_{jm}B_{jm}+\mathcal A_T$.

The stored `local_blocks[j, m]` is exactly $B_{jm}$.  The stored
`objective_terms$cell_terms` is $\sum_m\mathcal C_m$, and the stored assignment
information contains the expected log-assignment term, $H_Z$, and, where
applicable, the Dirichlet KL term.  `objective_history` records their sum at
the temperature in `temperature_history`.

## 12  Summary

The current fixed-$M$ method is characterized by six identities:

\[
q(C,Z,U)
=q(C)\prod_jq(Z_j)q(U_j\mid Z_j),
\]

\[
P_{jm}=\lambda_{jm}Q_K+D_{jm},
\qquad
P_{jm}\mu_{jm}=b_{jm},
\]

\[
\log r_{imk}
\stackrel{+c}{=}
\log\pi_k^{(m)}
+\sum_jw_{jm}
\mathbb E_{q_{jm}}[\log p(x_{ij}\mid C_i^{(m)}=k,U_j)],
\]

\[
\sigma_j^2=\frac1n\sum_mw_{jm}E_{jm},
\]

\[
w_{jm}\propto
\exp\{(B_{jm}+a_m)/T\},
\]

and

\[
\mathcal L_T
=\sum_m\mathcal C_m+\sum_{jm}w_{jm}B_{jm}+\mathcal A_T.
\]

Together these formulas explain why the implementation can use an intrinsic
random-walk prior without assignment-dependent precision regularization, why
the estimated variance is shared across orderings, and why convergence and
model selection must be interpreted within the fixed-$M$, fixed-temperature
setting for which the objective was defined.
